---
title: "Association Rule Mining (ARM)"
subtitle: "Introduction with R"
author: "Dr. Mustafa Hameed"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    show-slide-number: all
    progress: true
    chalkboard: false
    code-line-numbers: false
    highlight-style: github
    smaller: false
    scrollable: false
    transition: slide
    footer: "Association Rule Mining in R"
knitr:
  opts_chunk:
    echo: true
    warning: false
    message: false
---

## What is Association Rule Mining? {background-color="#2c3e50"}

::: incremental
-   Finds **interesting relationships** between items in a dataset
-   Most famous use case: **Market Basket Analysis**
-   Used in retail, healthcare, web mining, and recommendation systems
:::

. . .

> *"Customers who buy bread and butter also tend to buy milk."*
>
> `{bread, butter}` **→** `{milk}`

## Why Does It Matter?

::::: columns
::: {.column width="50%"}
**Retail**

-   Product placement
-   Bundle promotions
-   Cross-selling
:::

::: {.column width="50%"}
**Other Domains**

-   Medical co-diagnoses
-   Web click-stream patterns
-   Fraud detection
:::
:::::

## Three Key Measures {.smaller}

| Measure        | Formula               | What it tells us             |
|----------------|-----------------------|------------------------------|
| **Support**    | freq(A ∪ B) / N       | How *common* is the itemset? |
| **Confidence** | Supp(A ∪ B) / Supp(A) | How *reliable* is the rule?  |
| **Lift**       | Conf(A→B) / Supp(B)   | Is it *better than chance*?  |

. . .

::: callout-tip
## Rule of Thumb

**Lift \> 1** → useful rule · **Lift = 1** → no association · **Lift \< 1** → negative association
:::

## Support — Illustrated

**5 transactions:**

| TID | Items               |
|-----|---------------------|
| T1  | bread, butter, milk |
| T2  | bread, butter       |
| T3  | bread, milk         |
| T4  | butter, milk        |
| T5  | bread, butter, milk |

. . .

$$\text{Support}(\{bread, butter\}) = \frac{3}{5} = 0.60$$

## Confidence — Illustrated

Rule: **{bread, butter} → {milk}**

-   Transactions with {bread, butter}: T1, T2, T5 → **3**
-   Transactions with {bread, butter, **milk**}: T1, T5 → **2**

. . .

$$\text{Confidence} = \frac{2/5}{3/5} = \frac{2}{3} \approx 0.67$$

. . .

> 67% of the time when someone buys bread + butter, they also buy milk.

## Lift — Illustrated

$$\text{Lift} = \frac{\text{Confidence}}{\text{Support(milk)}} = \frac{0.67}{0.80} = 0.83$$

. . .

-   Lift \< 1 → buying bread & butter does **not** strongly predict milk here
-   We need **more data** and **lower thresholds** for small datasets

## The Apriori Algorithm

::: incremental
1.  **Set minimum support** — prune rare itemsets early
2.  **Find frequent 1-itemsets** `{bread}`, `{milk}`, …
3.  **Build frequent 2-itemsets** by joining 1-itemsets
4.  **Continue growing** until no new frequent itemsets are found
5.  **Generate rules** from frequent itemsets that meet min-confidence
:::

. . .

::: callout-note
The key insight: *a subset of a frequent itemset must also be frequent*\
(the **anti-monotone** property)
:::

## Step 1 — Load Packages

```{r}
#| label: setup
#| echo: true

if (!requireNamespace("arules",    quietly = TRUE)) install.packages("arules")
if (!requireNamespace("arulesViz", quietly = TRUE)) install.packages("arulesViz")

library(arules)
library(arulesViz)
library(ggplot2)

cat("Ready!\n")
```

## Step 2 — Create a Small Dataset

```{r}
#| label: dataset

baskets <- list(
  c("bread", "butter", "milk"),
  c("bread", "butter"),
  c("bread", "milk"),
  c("butter", "milk"),
  c("bread", "butter", "milk"),
  c("bread", "eggs"),
  c("bread", "butter", "eggs"),
  c("milk", "yogurt"),
  c("milk", "yogurt", "butter"),
  c("bread", "yogurt"),
  c("cheese", "bread", "butter"),
  c("cheese", "milk"),
  c("bread", "butter", "milk", "eggs"),
  c("yogurt", "cheese"),
  c("bread", "milk", "yogurt"),
  c("butter", "eggs"),
  c("bread", "cheese"),
  c("milk", "butter", "cheese"),
  c("bread", "butter", "yogurt"),
  c("eggs", "milk")
)

txns <- as(baskets, "transactions")
cat(length(txns), "transactions,", length(itemLabels(txns)), "items\n")
```

## Step 3 — Inspect the Data

```{r}
#| label: inspect

inspect(txns[1:5])
```

## Step 3 — Item Frequency Plot

```{r}
#| label: item-freq
#| fig-width: 7
#| fig-height: 4

itemFrequencyPlot(
  txns,
  type = "relative",
  col  = "steelblue",
  main = "Relative Item Frequency"
)
```

## Step 4 — Find Frequent Itemsets

```{r}
#| label: frequent-sets

frequent_sets <- apriori(
  txns,
  parameter = list(supp = 0.20, target = "frequent itemsets")
)

cat("Frequent itemsets:", length(frequent_sets), "\n\n")
inspect(sort(frequent_sets, by = "support")[1:8])
```

## Step 5 — Generate Rules

```{r}
#| label: rules

rules <- apriori(
  txns,
  parameter = list(supp = 0.20, conf = 0.50, minlen = 2)
)

cat("Rules generated:", length(rules), "\n")
inspect(sort(rules, by = "lift", decreasing = TRUE))
```

## Step 6 — Scatter Plot

```{r}
#| label: scatter
#| fig-width: 7
#| fig-height: 5

plot(rules, method = "scatterplot",
     measure = c("support", "confidence"),
     shading = "lift")
```

## Step 6 — Network Graph

```{r}
#| label: graph
#| fig-width: 8
#| fig-height: 5

plot(sort(rules, by = "lift"),
     method = "graph",
     main   = "Association Rules Network")
```

## Step 7 — Filter for a Specific Item

*What items predict buying **milk**?*

```{r}
#| label: target-milk

rules_milk <- apriori(
  txns,
  parameter  = list(supp = 0.15, conf = 0.50, minlen = 2),
  appearance = list(rhs = "milk", default = "lhs")
)

cat("Rules predicting milk:", length(rules_milk), "\n\n")
inspect(sort(rules_milk, by = "lift", decreasing = TRUE))
```

## Step 8 — Export to Data Frame

```{r}
#| label: dataframe

rules_df <- as(rules, "data.frame")
rules_df$support    <- round(rules_df$support,    3)
rules_df$confidence <- round(rules_df$confidence, 3)
rules_df$lift       <- round(rules_df$lift,       3)

head(rules_df[order(-rules_df$lift), ], 6)
```

## Summary {background-color="#2c3e50"}

| Step | What we did                                    |
|------|------------------------------------------------|
| 1    | Loaded `arules` and `arulesViz`                |
| 2    | Built a small transaction dataset              |
| 3    | Explored item frequencies                      |
| 4    | Found frequent itemsets with Apriori           |
| 5    | Generated rules (support + confidence filters) |
| 6    | Visualised rules (scatter & graph)             |
| 7    | Filtered rules for a specific product          |
| 8    | Exported rules to a data frame                 |

## Key Takeaways {background-color="#2c3e50"}

::: incremental
-   **Support** — how common is the pattern?
-   **Confidence** — how reliable is the rule?
-   **Lift \> 1** — the rule is better than guessing
-   **Apriori** is efficient because it prunes low-support candidates early
-   `arules` + `arulesViz` make ARM easy in R
:::

. . .

::: callout-tip
## Try it yourself!

Change `supp` and `conf` thresholds and see how the number of rules changes.
:::

## References

1.  Agrawal, R., & Srikant, R. (1994). *Fast algorithms for mining association rules*. VLDB.
2.  Hahsler, M., Grün, B., & Hornik, K. (2005). `arules` — A computational environment for mining association rules. *Journal of Statistical Software*, 14(15).
3.  Hahsler, M. (2017). `arulesViz`: Visualizing Association Rules and Frequent Itemsets. *Journal of Statistical Software*, 76(2).
