---
title: "Association Rule Mining in R — A Simple Introduction"
author: "Instructor"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
    theme: flatly
knitr:
  opts_chunk:
    echo: true
    warning: false
    message: false
---

## What is Association Rule Mining?

**Association Rule Mining (ARM)** finds interesting relationships between items
in a dataset. The classic example is **market basket analysis**:

> *"Customers who buy bread and butter also tend to buy milk."*

This gives us a rule:  `{bread, butter} → {milk}`

---

## The Three Key Measures

Before we write any code, let us understand the three numbers that describe
every rule.

### Support

How *common* is the itemset?

$$\text{Support}(A) = \frac{\text{Number of transactions containing } A}{\text{Total transactions}}$$

### Confidence

How *reliable* is the rule?

$$\text{Confidence}(A \Rightarrow B) = \frac{\text{Support}(A \cup B)}{\text{Support}(A)}$$

### Lift

Is the rule *better than random chance*?

$$\text{Lift}(A \Rightarrow B) = \frac{\text{Confidence}(A \Rightarrow B)}{\text{Support}(B)}$$

| Lift value | Meaning |
|-----------|---------|
| Lift > 1  | Positive association (useful rule) |
| Lift = 1  | No association (items are independent) |
| Lift < 1  | Negative association |

---

## A Tiny Example by Hand

Consider **5 transactions** in a small grocery shop:

| TID | Items bought |
|-----|-------------|
| T1  | bread, butter, milk |
| T2  | bread, butter |
| T3  | bread, milk |
| T4  | butter, milk |
| T5  | bread, butter, milk |

Let us calculate the metrics for the rule **{bread, butter} → {milk}**:

- Transactions containing {bread, butter}: T1, T2, T5 → **3**
- Transactions containing {bread, butter, milk}: T1, T5 → **2**
- Transactions containing {milk}: T1, T3, T4, T5 → **4**

$$\text{Support}(\{bread, butter, milk\}) = \frac{2}{5} = 0.40$$

$$\text{Confidence}(\{bread, butter\} \Rightarrow \{milk\}) = \frac{2/5}{3/5} = \frac{2}{3} \approx 0.67$$

$$\text{Lift} = \frac{0.67}{4/5} = \frac{0.67}{0.80} \approx 0.83$$

A lift of 0.83 (< 1) means buying bread and butter does **not** strongly predict
buying milk in this tiny dataset. We need more transactions to find meaningful
patterns!

---

## Step 1 — Install and Load Packages

We use two packages:

- **`arules`** — the core ARM engine
- **`arulesViz`** — visualisation tools

```{r}
#| label: setup

# Install packages if they are not already present
if (!requireNamespace("arules",     quietly = TRUE)) install.packages("arules")
if (!requireNamespace("arulesViz",  quietly = TRUE)) install.packages("arulesViz")
if (!requireNamespace("ggplot2",    quietly = TRUE)) install.packages("ggplot2")

library(arules)
library(arulesViz)
library(ggplot2)

cat("Packages loaded successfully.\n")
```

---

## Step 2 — Create a Small Dataset

We manually define **20 transactions** with only **8 grocery items**.
This keeps things easy to follow.

```{r}
#| label: dataset

# Each inner vector is one shopping basket
baskets <- list(
  c("bread", "butter", "milk"),
  c("bread", "butter"),
  c("bread", "milk"),
  c("butter", "milk"),
  c("bread", "butter", "milk"),
  c("bread", "eggs"),
  c("bread", "butter", "eggs"),
  c("milk", "yogurt"),
  c("milk", "yogurt", "butter"),
  c("bread", "yogurt"),
  c("cheese", "bread", "butter"),
  c("cheese", "milk"),
  c("bread", "butter", "milk", "eggs"),
  c("yogurt", "cheese"),
  c("bread", "milk", "yogurt"),
  c("butter", "eggs"),
  c("bread", "cheese"),
  c("milk", "butter", "cheese"),
  c("bread", "butter", "yogurt"),
  c("eggs", "milk")
)

# Convert to an arules 'transactions' object
txns <- as(baskets, "transactions")

cat("Number of transactions:", length(txns), "\n")
cat("Number of unique items:", length(itemLabels(txns)), "\n")
```

---

## Step 3 — Explore the Data

### Inspect the First Few Transactions

```{r}
#| label: inspect

inspect(txns[1:5])
```

### Summary of the Transaction Dataset

```{r}
#| label: summary-data

summary(txns)
```

### Item Frequency Plot

This shows how often each item appears across all transactions.

```{r}
#| label: item-freq
#| fig-cap: "Item frequency across all transactions"
#| fig-width: 7
#| fig-height: 4

itemFrequencyPlot(
  txns,
  type  = "relative",
  col   = "steelblue",
  main  = "Relative Item Frequency",
  ylab  = "Support (proportion of transactions)"
)
```

---

## Step 4 — Find Frequent Itemsets

The **Apriori algorithm** works in two stages:

1. Find all itemsets with support ≥ `min_support` (**frequent itemsets**)
2. Generate rules from those itemsets with confidence ≥ `min_confidence`

We start with Stage 1.

```{r}
#| label: frequent-itemsets

frequent_sets <- apriori(
  txns,
  parameter = list(
    supp   = 0.20,   # appear in at least 20% of transactions
    target = "frequent itemsets"
  )
)

cat("Total frequent itemsets found:", length(frequent_sets), "\n\n")

# Show the top 10 by support
inspect(sort(frequent_sets, by = "support", decreasing = TRUE)[1:10])
```

---

## Step 5 — Generate Association Rules

Now Stage 2 — turn frequent itemsets into rules.

```{r}
#| label: rules

rules <- apriori(
  txns,
  parameter = list(
    supp   = 0.20,   # minimum support
    conf   = 0.50,   # minimum confidence
    minlen = 2       # at least 2 items (antecedent + consequent)
  )
)

cat("Total rules generated:", length(rules), "\n")
```

---

## Step 6 — Inspect the Rules

### All Rules (sorted by Lift)

```{r}
#| label: view-rules

rules_sorted <- sort(rules, by = "lift", decreasing = TRUE)
inspect(rules_sorted)
```

### Top 5 Rules by Confidence

```{r}
#| label: top-conf

top_conf <- head(sort(rules, by = "confidence", decreasing = TRUE), 5)
inspect(top_conf)
```

---

## Step 7 — Visualise the Rules

### Scatter Plot (Support vs Confidence, coloured by Lift)

```{r}
#| label: scatter
#| fig-cap: "Scatter plot of all rules. Higher lift = stronger rule."
#| fig-width: 7
#| fig-height: 5

plot(
  rules,
  method  = "scatterplot",
  measure = c("support", "confidence"),
  shading = "lift"
)
```

### Network Graph of Top Rules

Each circle is an **item** and each arrow represents a rule.
The size of the arrow-node shows lift.

```{r}
#| label: graph
#| fig-cap: "Network graph of the top rules"
#| fig-width: 8
#| fig-height: 6

plot(
  rules_sorted,
  method = "graph",
  main   = "Association Rules Network"
)
```

---

## Step 8 — Filter Rules for Specific Items

In practice, you often want rules for a single product.

### What leads to buying *milk*?

```{r}
#| label: target-milk

rules_milk <- apriori(
  txns,
  parameter  = list(supp = 0.15, conf = 0.50, minlen = 2),
  appearance = list(rhs = "milk", default = "lhs")
)

cat("Rules predicting milk:", length(rules_milk), "\n\n")
inspect(sort(rules_milk, by = "lift", decreasing = TRUE))
```

### What leads to buying *yogurt*?

```{r}
#| label: target-yogurt

rules_yogurt <- apriori(
  txns,
  parameter  = list(supp = 0.10, conf = 0.40, minlen = 2),
  appearance = list(rhs = "yogurt", default = "lhs")
)

cat("Rules predicting yogurt:", length(rules_yogurt), "\n\n")
inspect(sort(rules_yogurt, by = "lift", decreasing = TRUE))
```

---

## Step 9 — Convert Rules to a Data Frame

You can export rules to a regular R data frame for further analysis.

```{r}
#| label: dataframe

rules_df <- as(rules, "data.frame")

# Round numeric columns for readability
rules_df$support    <- round(rules_df$support,    3)
rules_df$confidence <- round(rules_df$confidence, 3)
rules_df$lift       <- round(rules_df$lift,       3)

# Show all rules in a tidy table
rules_df[order(-rules_df$lift), ]
```

---

## Summary

In this lecture we covered:

| Step | Task |
|------|------|
| 1 | Install and load `arules` and `arulesViz` |
| 2 | Create / load a transaction dataset |
| 3 | Explore item frequencies |
| 4 | Find frequent itemsets with `apriori()` |
| 5 | Generate association rules |
| 6 | Inspect and sort rules by Support, Confidence, Lift |
| 7 | Visualise rules (scatter plot, network graph) |
| 8 | Filter rules for a specific item |
| 9 | Export rules to a data frame |

### Key Takeaways

- **Support** tells us how common an itemset is.
- **Confidence** tells us how often the rule is correct.
- **Lift > 1** means a genuinely useful association.
- The **Apriori algorithm** is the classic method to find these rules efficiently.

---

## References

1. Agrawal, R., & Srikant, R. (1994). *Fast algorithms for mining association rules*. VLDB.
2. Hahsler, M., Grün, B., & Hornik, K. (2005). `arules` — A computational environment for mining association rules. *Journal of Statistical Software*, 14(15).
3. Hahsler, M. (2017). `arulesViz`: Visualizing Association Rules and Frequent Itemsets. *Journal of Statistical Software*, 76(2).
