---
title: "Association Rule Mining (ARM) in R"
author: "Instructor"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
    fontsize: 11pt
    pdf-engine: lualatex
knitr:
  opts_chunk:
    echo: true
    warning: false
    message: false
---

## Introduction

**Association Rule Mining (ARM)** is an unsupervised machine learning technique used to
discover interesting relationships (associations) between variables in large datasets.
Common applications include:

- **Market basket analysis** (e.g., "customers who buy bread also buy butter")
- **Recommendation systems**
- **Medical diagnosis**
- **Web usage mining**

### Key Metrics

| Metric | Formula | Interpretation |
|--------|---------|----------------|
| **Support** | freq(A+B) / N | How often the itemset appears |
| **Confidence** | Support(A+B) / Support(A) | Reliability of the rule |
| **Lift** | Confidence(A=>B) / Support(B) | Strength vs. random chance |

A **Lift > 1** indicates that A and B appear together more often than expected by chance.

---

## Setup: Install and Load Packages

```{r}
#| label: setup

# Check that required packages are installed
required_pkgs <- c("arules", "arulesViz", "ggplot2", "reshape2", "RColorBrewer")
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, FUN.VALUE = logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  stop(
    "The following required packages are not installed: ",
    paste(missing_pkgs, collapse = ", "),
    "\nPlease install them before rendering this document, e.g.: ",
    'install.packages(c("arules", "arulesViz", "ggplot2", "reshape2", "RColorBrewer"))'
  )
}

library(arules)
library(arulesViz)
library(ggplot2)
library(reshape2)
library(RColorBrewer)

cat("All packages loaded successfully.\n")
```

---

## Dataset: Synthetic Grocery Transactions

We simulate a realistic market-basket dataset with **1,000 transactions** across
**20 grocery items**, then convert it to an `arules` `transactions` object.

```{r}
#| label: create-dataset

set.seed(42)

grocery_items <- c(
  "whole milk", "yogurt", "rolls/buns", "soda", "other vegetables",
  "root vegetables", "tropical fruit", "bottled water", "pastry",
  "citrus fruit", "shopping bags", "sausage", "newspapers",
  "bottled beer", "canned beer", "fruit/vegetable juice",
  "whipped/sour cream", "domestic eggs", "pip fruit", "chocolate"
)

# Co-purchase basket patterns
basket_patterns <- list(
  c("whole milk", "yogurt", "other vegetables"),
  c("whole milk", "rolls/buns", "soda"),
  c("root vegetables", "other vegetables", "whole milk"),
  c("tropical fruit", "yogurt", "whole milk"),
  c("pastry", "citrus fruit", "whole milk"),
  c("sausage", "rolls/buns", "soda"),
  c("bottled beer", "canned beer"),
  c("shopping bags", "other vegetables", "root vegetables"),
  c("domestic eggs", "whole milk", "other vegetables"),
  c("whipped/sour cream", "whole milk", "yogurt")
)

generate_basket <- function() {
  pattern <- basket_patterns[[sample(length(basket_patterns), 1)]]
  extra   <- sample(grocery_items, sample(0:2, 1), replace = FALSE)
  unique(c(pattern, extra))
}

basket_list <- replicate(1000, generate_basket(), simplify = FALSE)

# Convert to arules transactions object
txns <- as(basket_list, "transactions")

cat("Total transactions:", length(txns), "\n")
cat("Unique items:      ", length(itemLabels(txns)), "\n")
inspect(txns[1:3])
```

---

## Exploratory Data Analysis

### Dataset Summary

```{r}
#| label: dataset-summary

summary(txns)
```

### Top 15 Most Frequent Items

```{r}
#| label: item-frequency-abs
#| fig-cap: "Top 15 Most Frequent Items (Absolute Count)"
#| fig-width: 8
#| fig-height: 5

itemFrequencyPlot(
  txns,
  topN  = 15,
  type  = "absolute",
  col   = "steelblue",
  main  = "Top 15 Most Frequent Items",
  xlab  = "Items",
  ylab  = "Absolute Frequency"
)
```

### Top 15 Items by Relative Support

```{r}
#| label: item-frequency-rel
#| fig-cap: "Top 15 Items by Relative Support"
#| fig-width: 8
#| fig-height: 5

itemFrequencyPlot(
  txns,
  topN  = 15,
  type  = "relative",
  col   = "coral",
  main  = "Top 15 Items by Relative Support",
  xlab  = "Items",
  ylab  = "Support (Relative Frequency)"
)
```

### Item Frequency — ggplot2 Bar Chart

```{r}
#| label: item-frequency-gg
#| fig-cap: "Top 15 Items by Support (ggplot2)"
#| fig-width: 8
#| fig-height: 5

item_freq <- sort(itemFrequency(txns, type = "relative"), decreasing = TRUE)
top15_df  <- data.frame(
  Item    = names(item_freq[1:15]),
  Support = as.numeric(item_freq[1:15])
)

ggplot(top15_df, aes(x = reorder(Item, Support), y = Support)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 15 Grocery Items by Support",
    x     = "Item",
    y     = "Support"
  ) +
  theme_minimal(base_size = 11)
```

### Transaction Length Distribution

```{r}
#| label: tx-length
#| fig-cap: "Distribution of Transaction Lengths"
#| fig-width: 7
#| fig-height: 4

tx_sizes <- size(txns)

ggplot(data.frame(length = tx_sizes), aes(x = length)) +
  geom_histogram(binwidth = 1, fill = "coral", color = "white") +
  labs(
    title = "Transaction Length Distribution",
    x     = "Number of Items per Transaction",
    y     = "Frequency"
  ) +
  theme_minimal(base_size = 11)

cat(sprintf("Mean transaction length: %.2f\n", mean(tx_sizes)))
cat(sprintf("Max  transaction length: %d\n",   max(tx_sizes)))
```

---

## Mining Frequent Itemsets with Apriori

```{r}
#| label: apriori-itemsets

frequent_itemsets <- apriori(
  txns,
  parameter = list(
    supp   = 0.05,
    target = "frequent itemsets",
    minlen = 1,
    maxlen = 4
  )
)

cat("Total frequent itemsets:", length(frequent_itemsets), "\n")
inspect(sort(frequent_itemsets, by = "support", decreasing = TRUE)[1:10])
```

### Frequent Itemsets by Length

```{r}
#| label: itemsets-by-length
#| fig-cap: "Number of Frequent Itemsets by Length"
#| fig-width: 6
#| fig-height: 4

length_counts <- table(size(frequent_itemsets))
length_df     <- data.frame(
  Length = as.integer(names(length_counts)),
  Count  = as.integer(length_counts)
)

ggplot(length_df, aes(x = factor(Length), y = Count)) +
  geom_bar(stat = "identity", fill = "mediumseagreen", color = "white") +
  labs(
    title = "Frequent Itemsets by Length",
    x     = "Itemset Length",
    y     = "Count"
  ) +
  theme_minimal(base_size = 11)
```

---

## Generating Association Rules

```{r}
#| label: rules

rules <- apriori(
  txns,
  parameter = list(
    supp   = 0.05,
    conf   = 0.4,
    minlen = 2,
    maxlen = 4,
    target = "rules"
  )
)

cat("Total association rules:", length(rules), "\n")
rules_sorted <- sort(rules, by = "lift", decreasing = TRUE)
inspect(head(rules_sorted, 10))
```

---

## Inspecting Top Rules

### Top 10 Rules by Lift

```{r}
#| label: top-lift

top_lift <- head(sort(rules, by = "lift", decreasing = TRUE), 10)
inspect(top_lift)
```

### Top 10 Rules by Confidence

```{r}
#| label: top-confidence

top_conf <- head(sort(rules, by = "confidence", decreasing = TRUE), 10)
inspect(top_conf)
```

### Top 10 Rules by Support

```{r}
#| label: top-support

top_supp <- head(sort(rules, by = "support", decreasing = TRUE), 10)
inspect(top_supp)
```

---

## Visualizations

### Scatter Plot: Support vs Confidence (Lift as colour)

```{r}
#| label: scatter-supp-conf
#| fig-cap: "Scatter Plot: Support vs Confidence, coloured by Lift"
#| fig-width: 7
#| fig-height: 5

plot(
  rules,
  method  = "scatterplot",
  measure  = c("support", "confidence"),
  shading  = "lift",
  main     = "Association Rules: Support vs Confidence"
)
```

### Lift Distribution

```{r}
#| label: lift-dist
#| fig-cap: "Distribution of Lift Values Across All Rules"
#| fig-width: 7
#| fig-height: 4

lift_vals <- quality(rules)$lift
lift_df   <- data.frame(lift = lift_vals)

ggplot(lift_df, aes(x = lift)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 1.0, color = "red", linetype = "dashed",
             linewidth = 0.8) +
  annotate("text", x = 1.05, y = Inf, label = "Lift = 1",
           color = "red", hjust = 0, vjust = 2, size = 3.5) +
  labs(
    title = "Distribution of Lift Values",
    x     = "Lift",
    y     = "Number of Rules"
  ) +
  theme_minimal(base_size = 11)
```

### Confidence vs Lift (Support as colour)

```{r}
#| label: conf-lift
#| fig-cap: "Confidence vs Lift, coloured by Support"
#| fig-width: 7
#| fig-height: 5

rules_df <- as(rules, "data.frame")

ggplot(rules_df, aes(x = confidence, y = lift, colour = support)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_colour_viridis_c(name = "Support", option = "plasma") +
  labs(
    title = "Confidence vs Lift",
    x     = "Confidence",
    y     = "Lift"
  ) +
  theme_minimal(base_size = 11)
```

### Top 20 Rules — Graph-Based Network Plot

```{r}
#| label: graph-plot
#| fig-cap: "Graph of Top 20 Rules by Lift"
#| fig-width: 8
#| fig-height: 6

top20 <- head(sort(rules, by = "lift"), 20)
plot(
  top20,
  method = "graph",
  main   = "Top 20 Rules – Network Graph"
)
```

### Grouped Matrix Plot

```{r}
#| label: grouped-matrix
#| fig-cap: "Grouped Matrix of Top 50 Rules"
#| fig-width: 8
#| fig-height: 6

top50 <- head(sort(rules, by = "lift"), 50)
plot(
  top50,
  method = "grouped",
  main   = "Grouped Matrix of Top 50 Rules"
)
```

### Parallel Coordinates Plot

```{r}
#| label: paracoord
#| fig-cap: "Parallel Coordinates of Top 20 Rules"
#| fig-width: 8
#| fig-height: 5

plot(
  top20,
  method  = "paracoord",
  control = list(reorder = TRUE),
  main    = "Parallel Coordinates of Top 20 Rules"
)
```

### Top 15 Rules — Horizontal Bar Chart (Lift)

```{r}
#| label: bar-lift
#| fig-cap: "Top 15 Rules Ranked by Lift"
#| fig-width: 9
#| fig-height: 6

top15_rules <- head(sort(rules, by = "lift"), 15)
top15_df2   <- as(top15_rules, "data.frame")

# Format rule labels from the 'rules' column (e.g. "{a,b} => {c}")
top15_df2$rule  <- as.character(top15_df2$rules)
top15_df2$rule  <- factor(top15_df2$rule,
                           levels = top15_df2$rule[order(top15_df2$lift)])

ggplot(top15_df2, aes(x = rule, y = lift)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 15 Association Rules by Lift",
    x     = "Rule",
    y     = "Lift"
  ) +
  theme_minimal(base_size = 10)
```

---

## Targeted Rule Mining

### Rules with "whole milk" as Consequent

```{r}
#| label: targeted-milk

rules_milk <- apriori(
  txns,
  parameter  = list(supp = 0.03, conf = 0.4, minlen = 2),
  appearance = list(rhs = "whole milk", default = "lhs")
)
cat("Rules with whole milk as consequent:", length(rules_milk), "\n")
inspect(head(sort(rules_milk, by = "lift"), 10))
```

### Rules with "yogurt" as Consequent

```{r}
#| label: targeted-yogurt

rules_yogurt <- apriori(
  txns,
  parameter  = list(supp = 0.03, conf = 0.4, minlen = 2),
  appearance = list(rhs = "yogurt", default = "lhs")
)
cat("Rules with yogurt as consequent:", length(rules_yogurt), "\n")
inspect(head(sort(rules_yogurt, by = "lift"), 10))
```

---

## Removing Redundant Rules

```{r}
#| label: remove-redundant

non_redundant <- rules[!is.redundant(rules)]

cat("Total rules:         ", length(rules),        "\n")
cat("Non-redundant rules: ", length(non_redundant), "\n")
```

---

## Summary Statistics

```{r}
#| label: summary

cat("=== Association Rules Quality Summary ===\n")
print(summary(quality(rules)))
```

---

## Converting Rules to a Data Frame

```{r}
#| label: rules-dataframe

rules_df_full <- as(rules, "data.frame")
head(rules_df_full[order(-rules_df_full$lift), ], 10)
```

---

## Key Takeaways

- **Support** identifies how frequently an itemset appears in the dataset.
- **Confidence** measures the reliability of an association rule.
- **Lift > 1** indicates a positive association stronger than random chance.
- The **Apriori algorithm** prunes low-support candidates early, making it scalable.
- `arulesViz` provides rich visualization methods (scatter, graph, grouped matrix, parallel coordinates).

---

## References

1. Agrawal, R., & Srikant, R. (1994). *Fast algorithms for mining association rules*. VLDB.
2. Hahsler, M., Grün, B., & Hornik, K. (2005). `arules` — A computational environment for mining association rules. *Journal of Statistical Software*, 14(15).
3. Hahsler, M. (2017). `arulesViz`: Visualizing Association Rules and Frequent Itemsets. *Journal of Statistical Software*, 76(2).
