---
title: "Association Rule Mining in R — A Simple Introduction"
subtitle: "Computer Accessories Dataset"
author: "Dr. Mustafa Hameed"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: flatly
knitr:
  opts_chunk:
    echo: true
    warning: false
    message: false
---

## What is Association Rule Mining?

**Association Rule Mining (ARM)** is a data mining technique that discovers
*interesting relationships* (called **associations**) among items in large
datasets.

The classic application is **Market Basket Analysis** — analyzing which products
customers frequently purchase together.

> *"Customers who buy a laptop and a mouse also tend to buy a keyboard."*

This gives us a **rule**: `{laptop, mouse} → {keyboard}`

### Key Definitions

Before diving in, let us define the core vocabulary:

| Term | Definition |
|------|-----------|
| **Transaction** | A single purchase event (one receipt / one basket). Each transaction contains one or more items. |
| **Item** | An individual product in a transaction (e.g., "laptop", "mouse"). |
| **Itemset** | A collection of one or more items, e.g. `{laptop, mouse}`. |
| **Frequent Itemset** | An itemset whose support meets a minimum threshold. |
| **Association Rule** | An implication of the form `{A} → {B}`, read as "if A then B". |
| **Antecedent (LHS)** | The **left-hand side** — the "if" part of the rule, e.g. `{laptop, mouse}`. |
| **Consequent (RHS)** | The **right-hand side** — the "then" part of the rule, e.g. `{keyboard}`. |

---

## The Four Key Measures

Every association rule is evaluated by four important measures. We explain each
one with its formula, definition, and how to interpret it.

### Measure 1 — Support

> **Definition:** Support measures how *frequently* an itemset appears in the
> dataset. It tells us how **common** the pattern is.

$$\text{Support}(X) = \frac{\text{Number of transactions containing } X}{\text{Total number of transactions } (N)}$$

**How to interpret:**

- Support = 0.40 means the itemset appears in **40% of all transactions**.
- **Higher support** = more common pattern.
- We set a **minimum support threshold** to filter out rare, uninteresting patterns.
- Typical minimum: 0.01 to 0.30 depending on dataset size.

### Measure 2 — Confidence

> **Definition:** Confidence measures how *reliable* the rule is. Given that
> the antecedent (LHS) is purchased, how often is the consequent (RHS) also
> purchased?

$$\text{Confidence}(A \Rightarrow B) = \frac{\text{Support}(A \cup B)}{\text{Support}(A)}$$

**How to interpret:**

- Confidence = 0.80 means "**80% of the time** when A is bought, B is also bought."
- Ranges from 0 to 1 (i.e., 0% to 100%).
- Higher confidence = the rule fires correctly more often.
- ⚠️ **Caution:** A high confidence alone can be misleading if B is very popular (see Lift).

### Measure 3 — Lift

> **Definition:** Lift measures whether A and B appear together **more often
> than expected by chance**. It corrects for the popularity of B.

$$\text{Lift}(A \Rightarrow B) = \frac{\text{Confidence}(A \Rightarrow B)}{\text{Support}(B)}$$

**How to interpret:**

| Lift Value    | Meaning                                                      |
|--------------|--------------------------------------------------------------|
| **Lift > 1** | **Positive association** — items appear together more than expected. The rule is useful! |
| **Lift = 1** | **No association** — items are independent (appearing together is just by chance). |
| **Lift < 1** | **Negative association** — items tend to *avoid* each other.  |

**Rules of thumb:**

- Lift > 1.0 → worth investigating
- Lift > 1.5 → strong association
- Lift > 2.0 → very strong association

### Measure 4 — Conviction

> **Definition:** Conviction measures the degree of **dependence** between the
> antecedent and consequent. It compares the expected error rate (if items were
> independent) to the actual error rate of the rule.

$$\text{Conviction}(A \Rightarrow B) = \frac{1 - \text{Support}(B)}{1 - \text{Confidence}(A \Rightarrow B)}$$

**How to interpret:**

| Conviction Value | Meaning                                                      |
|-----------------|--------------------------------------------------------------|
| **Conv = 1**    | Items are **independent** — no dependency.                    |
| **Conv > 1**    | The rule is **meaningful** — items are dependent.             |
| **Conv → ∞**    | **Perfect rule** — confidence = 100%, the rule never fails.   |

**Rules of thumb:**

- Conviction > 1.0 → some dependency exists
- Conviction > 1.5 → strong dependency
- Conviction = ∞ → the rule has 100% confidence

---

## Our Dataset — Computer Accessories

We use a **simple, relatable dataset**: 10 transactions from a computer
accessories shop with 7 different items.

| TID | Items Purchased                       |
|-----|---------------------------------------|
| T1  | laptop, mouse, keyboard               |
| T2  | laptop, mouse                         |
| T3  | laptop, keyboard, USB hub             |
| T4  | mouse, keyboard, mousepad             |
| T5  | laptop, mouse, keyboard, USB hub      |
| T6  | mouse, mousepad                       |
| T7  | laptop, mouse, keyboard, webcam       |
| T8  | keyboard, headset                     |
| T9  | laptop, mouse, keyboard, mousepad     |
| T10 | laptop, keyboard                      |

---

## Hand Calculation — Counting Items

First, let us count how many transactions contain each item (**N = 10**):

| Item        | Transactions                    | Count | Support       |
|-------------|---------------------------------|-------|---------------|
| **laptop**  | T1, T2, T3, T5, T7, T9, T10    | **7** | 7/10 = **0.70** |
| **mouse**   | T1, T2, T4, T5, T6, T7, T9     | **7** | 7/10 = **0.70** |
| **keyboard**| T1, T3, T4, T5, T7, T8, T9, T10| **8** | 8/10 = **0.80** |
| **USB hub** | T3, T5                         | **2** | 2/10 = **0.20** |
| **mousepad**| T4, T6, T9                     | **3** | 3/10 = **0.30** |
| **webcam**  | T7                              | **1** | 1/10 = **0.10** |
| **headset** | T8                              | **1** | 1/10 = **0.10** |

---

## Hand Calculation — Rule 1: {laptop, mouse} → {keyboard}

Let us calculate **all four measures** step by step.

### Step 1 — Support

Find transactions containing **{laptop, mouse, keyboard}**:

- T1: laptop, mouse, keyboard ✓
- T5: laptop, mouse, keyboard, USB hub ✓
- T7: laptop, mouse, keyboard, webcam ✓
- T9: laptop, mouse, keyboard, mousepad ✓

Count = **4** → N = **10**

$$\boxed{\text{Support}(\{\text{laptop, mouse, keyboard}\}) = \frac{4}{10} = 0.40}$$

✅ **40% of all transactions** contain laptop + mouse + keyboard together.

### Step 2 — Confidence

Find transactions containing **{laptop, mouse}** (the antecedent):

- T1, T2, T5, T7, T9 → Count = **5** → Support = 5/10 = **0.50**

$$\text{Confidence} = \frac{\text{Support}(\{laptop, mouse, keyboard\})}{\text{Support}(\{laptop, mouse\})} = \frac{0.40}{0.50} = 0.80$$

✅ When a customer buys laptop + mouse, there is an **80% chance** they also buy a keyboard.

### Step 3 — Lift

Support({keyboard}) = 8/10 = **0.80**

$$\text{Lift} = \frac{0.80}{0.80} = 1.00$$

⚠️ Lift = 1.00 → buying laptop + mouse does **not** increase or decrease the chance of buying keyboard. They are **independent** at this level. This happens because keyboard is already in 80% of transactions!

### Step 4 — Conviction

$$\text{Conviction} = \frac{1 - \text{Support(keyboard)}}{1 - \text{Confidence}} = \frac{1 - 0.80}{1 - 0.80} = \frac{0.20}{0.20} = 1.00$$

⚠️ Conviction = 1.00 confirms **independence** — consistent with Lift = 1.00.

### Summary for Rule 1

| Measure        | Value | Interpretation                       |
|----------------|-------|--------------------------------------|
| **Support**    | 0.40  | Common — 40% of transactions         |
| **Confidence** | 0.80  | Very reliable — 80% accuracy         |
| **Lift**       | 1.00  | Independent — no better than chance   |
| **Conviction** | 1.00  | Confirms independence                 |

> **Key lesson:** A rule can have **high confidence but low lift**! The keyboard
> is so popular (80% support) that the high confidence is misleading. Always
> check Lift!

---

## Hand Calculation — Rule 2: {mouse} → {mousepad}

### Step 1 — Support

Transactions with **{mouse, mousepad}**: T4, T6, T9 → Count = **3**

$$\text{Support}(\{\text{mouse, mousepad}\}) = \frac{3}{10} = 0.30$$

### Step 2 — Confidence

Support({mouse}) = 7/10 = **0.70**

$$\text{Confidence} = \frac{0.30}{0.70} = 0.429$$

### Step 3 — Lift

Support({mousepad}) = 3/10 = **0.30**

$$\text{Lift} = \frac{0.429}{0.30} = 1.43$$

### Step 4 — Conviction

$$\text{Conviction} = \frac{1 - 0.30}{1 - 0.429} = \frac{0.70}{0.571} = 1.23$$

### Summary for Rule 2

| Measure        | Value | Interpretation                        |
|----------------|-------|---------------------------------------|
| **Support**    | 0.30  | Moderate — 30% of transactions        |
| **Confidence** | 0.429 | 43% of mouse buyers also buy mousepad |
| **Lift**       | 1.43  | **Positive** — 43% more likely than random  |
| **Conviction** | 1.23  | **Genuine dependency**                |

> ✅ **This is a useful rule!** A customer buying a mouse is 43% more likely to
> also buy a mousepad than a random customer. A shop could use this to create a
> bundle deal or place mousepads near mice.

---

## Hand Calculation — Rule 3: {laptop} → {mouse}

### Step 1 — Support

Transactions with **{laptop, mouse}**: T1, T2, T5, T7, T9 → Count = **5**

$$\text{Support} = \frac{5}{10} = 0.50$$

### Step 2 — Confidence

Support({laptop}) = 7/10 = **0.70**

$$\text{Confidence} = \frac{0.50}{0.70} = 0.714$$

### Step 3 — Lift

Support({mouse}) = 7/10 = **0.70**

$$\text{Lift} = \frac{0.714}{0.70} = 1.02$$

### Step 4 — Conviction

$$\text{Conviction} = \frac{1 - 0.70}{1 - 0.714} = \frac{0.30}{0.286} = 1.05$$

### Summary for Rule 3

| Measure        | Value | Interpretation                             |
|----------------|-------|--------------------------------------------|
| **Support**    | 0.50  | Very common — 50% of transactions          |
| **Confidence** | 0.714 | 71% of laptop buyers also buy a mouse      |
| **Lift**       | 1.02  | Very slightly positive — nearly independent |
| **Conviction** | 1.05  | Minimal dependency                          |

> This rule is **borderline**. The high support and confidence look appealing, but
> both laptop and mouse are individually very popular, so they naturally co-occur.
> Lift ≈ 1 confirms this.

---

## Comparing All Three Rules

| Rule | Support | Confidence | Lift | Conviction | Verdict |
|------|---------|-----------|------|-----------|---------|
| {laptop, mouse} → {keyboard} | 0.40 | 0.80 | **1.00** | 1.00 | Independent |
| {mouse} → {mousepad} | 0.30 | 0.43 | **1.43** | 1.23 | ✅ Useful |
| {laptop} → {mouse} | 0.50 | 0.71 | **1.02** | 1.05 | Borderline |

**Key insight:** The rule with the *highest confidence* (0.80) is actually the
*least useful* because its lift is only 1.00. The rule with the *highest lift*
(1.43) is the most actionable, even though its confidence is lower.

---

## How to Interpret Results — A Practical Guide

### Decision Framework

When you generate rules with R, use this checklist:

1. **Filter by Support** (≥ 0.10): Remove patterns too rare to be useful.
2. **Filter by Confidence** (≥ 0.50): Keep rules that are right at least half the time.
3. **Sort by Lift** (descending): The highest-lift rules are the most interesting.
4. **Check Conviction** (> 1): Confirm that the rule represents genuine dependency.

### Common Interpretation Pitfalls

| Pitfall | Example | Solution |
|---------|---------|----------|
| High confidence, low lift | Conf=0.80 but Lift=1.0 (popular consequent) | Always check Lift! |
| Very low support | Support=0.01 → only 1% of transactions | May be spurious — need more data |
| Lift < 1 | Items avoid each other | Still useful — avoid placing these items together |
| Ignoring conviction | Lift=1.2 looks good but Conv=1.01 | Weak dependency — be cautious |

### Business Actions Based on Results

| Lift Range | Action |
|-----------|--------|
| Lift > 2.0 | **Strong bundle** — create a package deal |
| Lift 1.5–2.0 | **Cross-sell** — recommend at checkout |
| Lift 1.1–1.5 | **Place nearby** — put on adjacent shelves |
| Lift ≈ 1.0 | **No action** — items are independent |
| Lift < 1.0 | **Separate** — avoid placing together |

---

## Step 1 — Install and Load Packages

We use two R packages:

- **`arules`** — the core ARM engine (implements the Apriori algorithm)
- **`arulesViz`** — visualisation tools for association rules

```{r}
#| label: setup

if (!requireNamespace("arules",     quietly = TRUE)) install.packages("arules")
if (!requireNamespace("arulesViz",  quietly = TRUE)) install.packages("arulesViz")
if (!requireNamespace("ggplot2",    quietly = TRUE)) install.packages("ggplot2")

library(arules)
library(arulesViz)
library(ggplot2)

cat("Packages loaded successfully.\n")
```

---

## Step 2 — Create the Computer Accessories Dataset

We manually define **10 transactions** with **7 items** — the same dataset
we used for hand calculations above.

```{r}
#| label: dataset

baskets <- list(
  c("laptop", "mouse", "keyboard"),           # T1
  c("laptop", "mouse"),                        # T2
  c("laptop", "keyboard", "USB hub"),          # T3
  c("mouse", "keyboard", "mousepad"),          # T4
  c("laptop", "mouse", "keyboard", "USB hub"), # T5
  c("mouse", "mousepad"),                      # T6
  c("laptop", "mouse", "keyboard", "webcam"),  # T7
  c("keyboard", "headset"),                    # T8
  c("laptop", "mouse", "keyboard", "mousepad"),# T9
  c("laptop", "keyboard")                      # T10
)

txns <- as(baskets, "transactions")

cat("Number of transactions:", length(txns), "\n")
cat("Number of unique items:", length(itemLabels(txns)), "\n")
cat("Items:", paste(itemLabels(txns), collapse = ", "), "\n")
```

---

## Step 3 — Explore the Data

### Inspect All Transactions

```{r}
#| label: inspect

inspect(txns)
```

### Summary of the Transaction Dataset

```{r}
#| label: summary-data

summary(txns)
```

### Item Frequency Plot

This bar chart shows how often each item appears across all transactions
(i.e., the **support** of each individual item).

```{r}
#| label: item-freq
#| fig-cap: "Item frequency (support) across all transactions"
#| fig-width: 7
#| fig-height: 4

itemFrequencyPlot(
  txns,
  type  = "relative",
  col   = "steelblue",
  main  = "Item Frequency — Computer Accessories",
  ylab  = "Support (proportion of transactions)"
)
```

---

## Step 4 — Verify Hand Calculations in R

Let us use R to confirm the numbers we calculated by hand.

### Rule 1: {laptop, mouse} → {keyboard}

```{r}
#| label: verify-rule1

cat("=== Rule 1: {laptop, mouse} -> {keyboard} ===\n\n")

supp_lmk <- sum(sapply(baskets, function(b) all(c("laptop","mouse","keyboard") %in% b))) / 10
supp_lm  <- sum(sapply(baskets, function(b) all(c("laptop","mouse") %in% b))) / 10
supp_k   <- sum(sapply(baskets, function(b) "keyboard" %in% b)) / 10

conf1 <- supp_lmk / supp_lm
lift1 <- conf1 / supp_k
conv1 <- (1 - supp_k) / (1 - conf1)

cat("Support({laptop,mouse,keyboard}) =", supp_lmk, "\n")
cat("Support({laptop,mouse})          =", supp_lm,  "\n")
cat("Support({keyboard})              =", supp_k,   "\n\n")
cat("Confidence =", round(conf1, 4), "\n")
cat("Lift       =", round(lift1, 4), "\n")
cat("Conviction =", round(conv1, 4), "\n")
```

### Rule 2: {mouse} → {mousepad}

```{r}
#| label: verify-rule2

cat("=== Rule 2: {mouse} -> {mousepad} ===\n\n")

supp_mm  <- sum(sapply(baskets, function(b) all(c("mouse","mousepad") %in% b))) / 10
supp_m   <- sum(sapply(baskets, function(b) "mouse" %in% b)) / 10
supp_mp  <- sum(sapply(baskets, function(b) "mousepad" %in% b)) / 10

conf2 <- supp_mm / supp_m
lift2 <- conf2 / supp_mp
conv2 <- (1 - supp_mp) / (1 - conf2)

cat("Support({mouse,mousepad}) =", supp_mm, "\n")
cat("Support({mouse})          =", supp_m,  "\n")
cat("Support({mousepad})       =", supp_mp, "\n\n")
cat("Confidence =", round(conf2, 4), "\n")
cat("Lift       =", round(lift2, 4), "\n")
cat("Conviction =", round(conv2, 4), "\n")
```

### Rule 3: {laptop} → {mouse}

```{r}
#| label: verify-rule3

cat("=== Rule 3: {laptop} -> {mouse} ===\n\n")

supp_lm2  <- sum(sapply(baskets, function(b) all(c("laptop","mouse") %in% b))) / 10
supp_l    <- sum(sapply(baskets, function(b) "laptop" %in% b)) / 10
supp_m2   <- sum(sapply(baskets, function(b) "mouse" %in% b)) / 10

conf3 <- supp_lm2 / supp_l
lift3 <- conf3 / supp_m2
conv3 <- (1 - supp_m2) / (1 - conf3)

cat("Support({laptop,mouse}) =", supp_lm2, "\n")
cat("Support({laptop})       =", supp_l,   "\n")
cat("Support({mouse})        =", supp_m2,  "\n\n")
cat("Confidence =", round(conf3, 4), "\n")
cat("Lift       =", round(lift3, 4), "\n")
cat("Conviction =", round(conv3, 4), "\n")
```

---

## Step 5 — Find Frequent Itemsets

The **Apriori algorithm** works in two stages:

1. Find all itemsets with support ≥ `min_support` (**frequent itemsets**)
2. Generate rules from those itemsets with confidence ≥ `min_confidence`

We start with Stage 1.

```{r}
#| label: frequent-itemsets

frequent_sets <- apriori(
  txns,
  parameter = list(
    supp   = 0.20,   # appear in at least 20% of transactions (2 out of 10)
    target = "frequent itemsets"
  )
)

cat("Total frequent itemsets found:", length(frequent_sets), "\n\n")
inspect(sort(frequent_sets, by = "support", decreasing = TRUE))
```

---

## Step 6 — Generate Association Rules

Now Stage 2 — turn frequent itemsets into rules.

```{r}
#| label: rules

rules <- apriori(
  txns,
  parameter = list(
    supp   = 0.20,   # minimum support
    conf   = 0.50,   # minimum confidence
    minlen = 2       # at least 2 items (antecedent + consequent)
  )
)

cat("Total rules generated:", length(rules), "\n")
```

---

## Step 7 — Inspect the Rules

### All Rules (sorted by Lift)

```{r}
#| label: view-rules

rules_sorted <- sort(rules, by = "lift", decreasing = TRUE)
inspect(rules_sorted)
```

### How to Read the Output

| Column | Meaning |
|--------|---------|
| **lhs** | Left-hand side (antecedent) — the "if" items |
| **rhs** | Right-hand side (consequent) — the "then" item |
| **support** | Fraction of all transactions containing both sides together |
| **confidence** | Of transactions containing lhs, what fraction also contains rhs? |
| **coverage** | Support of lhs alone (how common is the antecedent?) |
| **lift** | > 1 = positive association · = 1 = independent · < 1 = negative |
| **count** | Absolute number of transactions containing the full rule |

### Top 5 Rules by Confidence

```{r}
#| label: top-conf

top_conf <- head(sort(rules, by = "confidence", decreasing = TRUE), 5)
inspect(top_conf)
```

> Notice how the highest-confidence rules may not have the highest lift. 
> Always **sort by lift** to find the most interesting rules.

---

## Step 8 — Visualise the Rules

### Scatter Plot (Support vs Confidence, coloured by Lift)

Each dot is one rule. Look for rules in the **upper-right corner** (high
support AND high confidence) with **dark shading** (high lift).

```{r}
#| label: scatter
#| fig-cap: "Scatter plot of all rules. Darker = higher lift = stronger rule."
#| fig-width: 7
#| fig-height: 5

plot(
  rules,
  method  = "scatterplot",
  measure = c("support", "confidence"),
  shading = "lift",
  main    = "Rules: Support vs Confidence (shaded by Lift)"
)
```

### Network Graph of Top Rules

Each circle is an **item** and each arrow represents a rule.
Larger/darker nodes indicate higher lift.

```{r}
#| label: graph
#| fig-cap: "Network graph of top rules — Computer Accessories"
#| fig-width: 8
#| fig-height: 6

plot(
  rules_sorted,
  method = "graph",
  main   = "Association Rules — Computer Accessories"
)
```

---

## Step 9 — Filter Rules for Specific Items

In practice, you often want rules for a **specific product** — e.g., "what
items lead customers to also buy a keyboard?"

### What leads to buying a *keyboard*?

```{r}
#| label: target-keyboard

rules_keyboard <- apriori(
  txns,
  parameter  = list(supp = 0.15, conf = 0.50, minlen = 2),
  appearance = list(rhs = "keyboard", default = "lhs")
)

cat("Rules predicting keyboard:", length(rules_keyboard), "\n\n")
inspect(sort(rules_keyboard, by = "lift", decreasing = TRUE))
```

### What leads to buying a *mousepad*?

```{r}
#| label: target-mousepad

rules_mousepad <- apriori(
  txns,
  parameter  = list(supp = 0.10, conf = 0.30, minlen = 2),
  appearance = list(rhs = "mousepad", default = "lhs")
)

cat("Rules predicting mousepad:", length(rules_mousepad), "\n\n")
inspect(sort(rules_mousepad, by = "lift", decreasing = TRUE))
```

---

## Step 10 — Convert Rules to a Data Frame

You can export rules to a regular R data frame for further analysis or export.

```{r}
#| label: dataframe

rules_df <- as(rules, "data.frame")

# Round numeric columns for readability
rules_df$support    <- round(rules_df$support,    3)
rules_df$confidence <- round(rules_df$confidence, 3)
rules_df$lift       <- round(rules_df$lift,       3)

# Show all rules sorted by lift
rules_df[order(-rules_df$lift), ]
```

---

## Summary

In this lecture we covered:

| Step | Task |
|------|------|
| 1 | Install and load `arules` and `arulesViz` |
| 2 | Create a computer accessories transaction dataset |
| 3 | Explore item frequencies |
| 4 | Verify hand calculations of all four measures in R |
| 5 | Find frequent itemsets with `apriori()` |
| 6 | Generate association rules |
| 7 | Inspect and sort rules by Support, Confidence, Lift |
| 8 | Visualise rules (scatter plot, network graph) |
| 9 | Filter rules for specific items |
| 10 | Export rules to a data frame |

### The Four Measures — Final Summary

| Measure | Formula | What it tells us | Good values |
|---------|---------|-----------------|-------------|
| **Support** | freq(A∪B) / N | How common is the pattern? | ≥ 0.10 |
| **Confidence** | Supp(A∪B) / Supp(A) | How reliable is the rule? | ≥ 0.50 |
| **Lift** | Conf / Supp(B) | Better than random chance? | **> 1** (higher = better) |
| **Conviction** | (1−Supp(B)) / (1−Conf) | Are items dependent? | **> 1** (higher = stronger) |

### Key Takeaways

- **Support** tells us how common an itemset is.
- **Confidence** tells us how often the rule is correct.
- **Lift > 1** means a genuinely useful association (better than random).
- **Conviction > 1** means items are truly dependent.
- ⚠️ **Always check Lift** — high confidence can be misleading if the consequent is popular!
- The **Apriori algorithm** efficiently finds frequent itemsets by pruning using the anti-monotone property.

---

## References

1. Agrawal, R., & Srikant, R. (1994). *Fast algorithms for mining association rules*. VLDB.
2. Hahsler, M., Grün, B., & Hornik, K. (2005). `arules` — A computational environment for mining association rules. *Journal of Statistical Software*, 14(15).
3. Hahsler, M. (2017). `arulesViz`: Visualizing Association Rules and Frequent Itemsets. *Journal of Statistical Software*, 76(2).
4. Brin, S., Motwani, R., Ullman, J., & Tsur, S. (1997). *Dynamic itemset counting and implication rules for market basket data*. ACM SIGMOD.
