---
title: "Association Rule Mining (ARM)"
subtitle: "Introduction with R — Computer Accessories Dataset"
author: "Dr. Mustafa Hameed"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    show-slide-number: all
    progress: true
    chalkboard: false
    code-line-numbers: false
    highlight-style: github
    smaller: false
    scrollable: true
    transition: slide
    footer: "Association Rule Mining in R"
knitr:
  opts_chunk:
    echo: true
    warning: false
    message: false
---

## What is Association Rule Mining? {background-color="#2c3e50"}

::: {.callout-note appearance="minimal"}
**Association Rule Mining (ARM)** is a data mining technique that discovers
*interesting relationships* (associations) among items in large datasets.
:::

::: incremental
-   Finds **hidden patterns** — which items are purchased together?
-   Classic use case: **Market Basket Analysis**
-   Applications: retail, e-commerce, healthcare, web mining
:::

. . .

> *"Customers who buy a laptop and a mouse also tend to buy a keyboard."*
>
> `{laptop, mouse}` **→** `{keyboard}`

## Why Does ARM Matter?

::::: columns
::: {.column width="50%"}
**Business Uses**

-   **Product placement** — put associated items nearby
-   **Bundle deals** — "buy laptop + mouse, get keyboard 20% off"
-   **Cross-selling** — recommend items at checkout
-   **Inventory** — stock associated items together
:::

::: {.column width="50%"}
**Other Domains**

-   Medical co-diagnoses
-   Web click-stream patterns
-   Course enrollment patterns
-   Fraud detection
:::
:::::

## Key Definitions {background-color="#2c3e50"}

| Term | Definition |
|------|-----------|
| **Transaction** | A single purchase event (one receipt / one basket) |
| **Item** | An individual product in a transaction |
| **Itemset** | A collection of one or more items, e.g. `{laptop, mouse}` |
| **Association Rule** | An implication: `{A} → {B}` ("if A then B") |
| **Antecedent (LHS)** | The "if" part — `{laptop, mouse}` |
| **Consequent (RHS)** | The "then" part — `{keyboard}` |

## The Four Key Measures {.smaller}

| # | Measure        | Formula                        | Question it answers               |
|---|---------------|--------------------------------|-----------------------------------|
| 1 | **Support**    | freq(A ∪ B) / N               | How *common* is this itemset?     |
| 2 | **Confidence** | Support(A ∪ B) / Support(A)   | How *reliable* is this rule?      |
| 3 | **Lift**       | Confidence(A→B) / Support(B)  | Is it *better than random chance*?|
| 4 | **Conviction** | (1 − Support(B)) / (1 − Conf) | How *dependent* is the rule?      |

. . .

::: callout-tip
## Quick Guide

-   **Support** = frequency · **Confidence** = reliability · **Lift** = strength · **Conviction** = dependency
:::

## Our Dataset — Computer Accessories {background-color="#2c3e50"}

**10 transactions** from a computer accessories shop:

| TID | Items Purchased                       |
|-----|---------------------------------------|
| T1  | laptop, mouse, keyboard               |
| T2  | laptop, mouse                         |
| T3  | laptop, keyboard, USB hub             |
| T4  | mouse, keyboard, mousepad             |
| T5  | laptop, mouse, keyboard, USB hub      |
| T6  | mouse, mousepad                       |
| T7  | laptop, mouse, keyboard, webcam       |
| T8  | keyboard, headset                     |
| T9  | laptop, mouse, keyboard, mousepad     |
| T10 | laptop, keyboard                      |

## Item Counts (by hand) {.smaller}

Count how often each item appears **(N = 10)**:

| Item       | Transactions               | Count | Support      |
|------------|---------------------------|-------|--------------|
| **laptop**    | T1,T2,T3,T5,T7,T9,T10       | **7** | 7/10 = 0.70 |
| **mouse**     | T1,T2,T4,T5,T6,T7,T9        | **7** | 7/10 = 0.70 |
| **keyboard**  | T1,T3,T4,T5,T7,T8,T9,T10    | **8** | 8/10 = 0.80 |
| **USB hub**   | T3, T5                      | **2** | 2/10 = 0.20 |
| **mousepad**  | T4, T6, T9                  | **3** | 3/10 = 0.30 |
| **webcam**    | T7                          | **1** | 1/10 = 0.10 |
| **headset**   | T8                          | **1** | 1/10 = 0.10 |

## Measure 1 — Support {background-color="#34495e"}

::: {.callout-note}
## Definition
**Support** measures how frequently an itemset appears in the dataset.

$$\text{Support}(X) = \frac{\text{Number of transactions containing } X}{\text{Total number of transactions } (N)}$$
:::

-   Higher support = the pattern is more *common*
-   A support of 0.50 means the itemset appears in 50% of all transactions
-   Used to **filter out rare patterns** (set a minimum threshold)

## Support — Step-by-Step Calculation

**Rule: {laptop, mouse} → {keyboard}**

**Step 1:** Find transactions containing `{laptop, mouse, keyboard}`:

-   T1: laptop, mouse, keyboard ✓
-   T5: laptop, mouse, keyboard, USB hub ✓
-   T7: laptop, mouse, keyboard, webcam ✓
-   T9: laptop, mouse, keyboard, mousepad ✓

. . .

**Step 2:** Count = **4** out of N = **10**

$$\boxed{\text{Support}(\{\text{laptop, mouse, keyboard}\}) = \frac{4}{10} = 0.40}$$

> ✅ **Interpretation:** 40% of all transactions contain laptop + mouse + keyboard together.

## Measure 2 — Confidence {background-color="#34495e"}

::: {.callout-note}
## Definition
**Confidence** measures how often the rule is correct — given the antecedent, how often does the consequent appear?

$$\text{Confidence}(A \Rightarrow B) = \frac{\text{Support}(A \cup B)}{\text{Support}(A)}$$
:::

-   Confidence of 0.80 = "80% of the time when A is bought, B is also bought"
-   Ranges from 0 to 1 (0% to 100%)
-   **Higher is better**, but always verify with Lift!

## Confidence — Step-by-Step Calculation

**Rule: {laptop, mouse} → {keyboard}**

**Step 1:** Support({laptop, mouse, keyboard}) = 4/10 = **0.40**

**Step 2:** Transactions with {laptop, mouse}: T1, T2, T5, T7, T9 → **5**

-   Support({laptop, mouse}) = 5/10 = **0.50**

. . .

**Step 3:** Apply the formula:

$$\text{Confidence} = \frac{0.40}{0.50} = 0.80$$

. . .

> ✅ **Interpretation:** When a customer buys a laptop and mouse, there is an **80% chance** they also buy a keyboard.

## Measure 3 — Lift {background-color="#34495e"}

::: {.callout-note}
## Definition
**Lift** measures whether A and B appear together *more often than expected by chance*.

$$\text{Lift}(A \Rightarrow B) = \frac{\text{Confidence}(A \Rightarrow B)}{\text{Support}(B)}$$
:::

| Lift Value    | Meaning                                               |
|--------------|-------------------------------------------------------|
| **Lift > 1** | Positive association — items appear together more than expected |
| **Lift = 1** | No association — items are independent                 |
| **Lift < 1** | Negative association — items *avoid* each other        |

## Lift — Step-by-Step Calculation

**Rule: {laptop, mouse} → {keyboard}**

**Step 1:** Confidence = **0.80**

**Step 2:** Support({keyboard}) = 8/10 = **0.80**

. . .

**Step 3:** Apply the formula:

$$\text{Lift} = \frac{0.80}{0.80} = 1.00$$

. . .

> ⚠️ **Interpretation:** Lift = 1.00 means laptop + mouse does **not** increase or decrease the chance of buying a keyboard — they are **independent**.

## Measure 4 — Conviction {background-color="#34495e"}

::: {.callout-note}
## Definition
**Conviction** measures the degree of dependence between antecedent and consequent. It compares the expected error rate (if independent) to the actual error rate.

$$\text{Conviction}(A \Rightarrow B) = \frac{1 - \text{Support}(B)}{1 - \text{Confidence}(A \Rightarrow B)}$$
:::

| Conviction    | Meaning                                        |
|--------------|-------------------------------------------------|
| **Conv = 1** | Items are independent (no dependency)            |
| **Conv > 1** | The rule is meaningful — items are dependent     |
| **Conv → ∞** | Perfect rule (confidence = 100%)                 |

## Conviction — Step-by-Step Calculation

**Rule: {laptop, mouse} → {keyboard}**

**Step 1:** 1 − Support({keyboard}) = 1 − 0.80 = **0.20**

**Step 2:** 1 − Confidence = 1 − 0.80 = **0.20**

. . .

**Step 3:** Apply the formula:

$$\text{Conviction} = \frac{0.20}{0.20} = 1.00$$

. . .

> ⚠️ **Interpretation:** Conviction = 1.00 confirms **independence** — consistent with Lift = 1.00.

## Another Rule: {mouse} → {mousepad} {background-color="#2c3e50"}

| Step | Calculation | Result |
|------|------------|--------|
| Txns with {mouse, mousepad} | T4, T6, T9 | **3** |
| Support({mouse, mousepad}) | 3/10 | **0.30** |
| Support({mouse}) | 7/10 | **0.70** |
| Support({mousepad}) | 3/10 | **0.30** |
| **Confidence** | 0.30 / 0.70 | **0.429** |
| **Lift** | 0.429 / 0.30 | **1.43** |
| **Conviction** | (1−0.30) / (1−0.429) | **1.23** |

. . .

> ✅ **Lift = 1.43** and **Conviction = 1.23** — a meaningful positive association!
> Buying a mouse **increases** the chance of buying a mousepad by 43% over random.

## Another Rule: {laptop} → {mouse} {background-color="#2c3e50"}

| Step | Calculation | Result |
|------|------------|--------|
| Txns with {laptop, mouse} | T1,T2,T5,T7,T9 | **5** |
| Support({laptop, mouse}) | 5/10 | **0.50** |
| Support({laptop}) | 7/10 | **0.70** |
| Support({mouse}) | 7/10 | **0.70** |
| **Confidence** | 0.50 / 0.70 | **0.714** |
| **Lift** | 0.714 / 0.70 | **1.02** |
| **Conviction** | (1−0.70)/(1−0.714) | **1.05** |

. . .

> Lift = 1.02, Conviction = 1.05 — very *slight* positive association. Nearly independent.

## How to Interpret Results {background-color="#2c3e50"}

| Measure    | What to look for                                           |
|-----------|-----------------------------------------------------------|
| **Support**    | Filter out rare patterns (min 0.10–0.30)                    |
| **Confidence** | High confidence (≥ 0.50) = rule fires correctly often       |
| **Lift**       | **> 1** = useful · **> 1.5** = strong · **> 2** = very strong |
| **Conviction** | **> 1** = dependent · **> 1.5** = strong dependency          |

. . .

::: callout-important
## Watch Out!
**High confidence + low lift** → the consequent is just very popular!
Example: confidence({laptop,mouse}→{keyboard}) = 0.80, but lift = 1.0 because keyboard is in 80% of transactions.

**Always check Lift and Conviction** before acting on a rule!
:::

## The Apriori Algorithm

::: incremental
1.  **Set minimum support** — prune rare itemsets early
2.  **Find frequent 1-itemsets** — `{laptop}`, `{mouse}`, `{keyboard}`, …
3.  **Build frequent 2-itemsets** by joining 1-itemsets
4.  **Continue growing** until no new frequent itemsets found
5.  **Generate rules** that meet minimum confidence
:::

. . .

::: callout-note
**Anti-monotone property:** *A subset of a frequent itemset must also be frequent.*
If `{laptop, mouse}` is not frequent, then `{laptop, mouse, keyboard}` cannot be frequent — so we **prune** it!
:::

## Step 1 — Load Packages

```{r}
#| label: setup
#| echo: true

if (!requireNamespace("arules",    quietly = TRUE)) install.packages("arules")
if (!requireNamespace("arulesViz", quietly = TRUE)) install.packages("arulesViz")

library(arules)
library(arulesViz)
library(ggplot2)

cat("Ready!\n")
```

## Step 2 — Computer Accessories Dataset

```{r}
#| label: dataset

baskets <- list(
  c("laptop", "mouse", "keyboard"),
  c("laptop", "mouse"),
  c("laptop", "keyboard", "USB hub"),
  c("mouse", "keyboard", "mousepad"),
  c("laptop", "mouse", "keyboard", "USB hub"),
  c("mouse", "mousepad"),
  c("laptop", "mouse", "keyboard", "webcam"),
  c("keyboard", "headset"),
  c("laptop", "mouse", "keyboard", "mousepad"),
  c("laptop", "keyboard")
)

txns <- as(baskets, "transactions")
cat(length(txns), "transactions,", length(itemLabels(txns)), "unique items\n")
cat("Items:", paste(itemLabels(txns), collapse = ", "), "\n")
```

## Step 3 — Inspect the Data

```{r}
#| label: inspect

inspect(txns)
```

## Step 3b — Item Frequency Plot

```{r}
#| label: item-freq
#| fig-width: 7
#| fig-height: 4

itemFrequencyPlot(
  txns,
  type = "relative",
  col  = "steelblue",
  main = "Item Frequency — Computer Accessories",
  ylab = "Support (proportion of transactions)"
)
```

## Step 4 — Verify Hand Calculations in R

```{r}
#| label: verify-hand

cat("=== Verifying: {laptop, mouse} -> {keyboard} ===\n\n")

supp_lmk <- sum(sapply(baskets, function(b) all(c("laptop","mouse","keyboard") %in% b))) / 10
supp_lm  <- sum(sapply(baskets, function(b) all(c("laptop","mouse") %in% b))) / 10
supp_k   <- sum(sapply(baskets, function(b) "keyboard" %in% b)) / 10

conf <- supp_lmk / supp_lm
lift <- conf / supp_k
conv <- (1 - supp_k) / (1 - conf)

cat("Support({laptop,mouse,keyboard}) =", supp_lmk, "\n")
cat("Support({laptop,mouse})          =", supp_lm, "\n")
cat("Support({keyboard})              =", supp_k, "\n\n")
cat("Confidence =", round(conf, 4), "\n")
cat("Lift       =", round(lift, 4), "\n")
cat("Conviction =", round(conv, 4), "\n")
```

## Step 4b — Verify Second Rule in R

```{r}
#| label: verify-hand-2

cat("=== Verifying: {mouse} -> {mousepad} ===\n\n")

supp_mm  <- sum(sapply(baskets, function(b) all(c("mouse","mousepad") %in% b))) / 10
supp_m   <- sum(sapply(baskets, function(b) "mouse" %in% b)) / 10
supp_mp  <- sum(sapply(baskets, function(b) "mousepad" %in% b)) / 10

conf2 <- supp_mm / supp_m
lift2 <- conf2 / supp_mp
conv2 <- (1 - supp_mp) / (1 - conf2)

cat("Support({mouse,mousepad}) =", supp_mm, "\n")
cat("Support({mouse})          =", supp_m, "\n")
cat("Support({mousepad})       =", supp_mp, "\n\n")
cat("Confidence =", round(conf2, 4), "\n")
cat("Lift       =", round(lift2, 4), "\n")
cat("Conviction =", round(conv2, 4), "\n")
```

## Step 5 — Find Frequent Itemsets

```{r}
#| label: frequent-sets

frequent_sets <- apriori(
  txns,
  parameter = list(supp = 0.20, target = "frequent itemsets")
)

cat("Frequent itemsets found:", length(frequent_sets), "\n\n")
inspect(sort(frequent_sets, by = "support", decreasing = TRUE))
```

## Step 6 — Generate Association Rules

```{r}
#| label: rules

rules <- apriori(
  txns,
  parameter = list(supp = 0.20, conf = 0.50, minlen = 2)
)

cat("Rules generated:", length(rules), "\n\n")
inspect(sort(rules, by = "lift", decreasing = TRUE))
```

## Step 7 — Interpreting R Output {.smaller}

| Column     | What it means                                                |
|------------|--------------------------------------------------------------|
| **lhs**        | Left-hand side (antecedent) — the "if" items                |
| **rhs**        | Right-hand side (consequent) — the "then" item               |
| **support**    | Fraction of all transactions containing both sides           |
| **confidence** | Of transactions with lhs, what fraction also has rhs?        |
| **coverage**   | Support of lhs alone (how common is the antecedent?)         |
| **lift**       | > 1 = positive association · = 1 = independent · < 1 = negative |
| **count**      | Absolute number of transactions containing the full rule     |

::: callout-tip
## Reading a Rule
`{laptop, mouse} → {keyboard} [supp=0.40, conf=0.80, lift=1.00]`

*"40% of transactions have all three. When laptop + mouse are bought, 80% also buy keyboard. But lift=1 — this is no better than chance because keyboard is already very popular."*
:::

## Step 8 — Scatter Plot

```{r}
#| label: scatter
#| fig-width: 7
#| fig-height: 5

plot(rules, method = "scatterplot",
     measure = c("support", "confidence"),
     shading = "lift",
     main = "Rules: Support vs Confidence (shaded by Lift)")
```

## Step 8b — Network Graph

```{r}
#| label: graph
#| fig-width: 8
#| fig-height: 5

plot(sort(rules, by = "lift"),
     method = "graph",
     main   = "Association Rules — Computer Accessories")
```

## Step 9 — Filter for a Specific Item

*What items predict buying a **keyboard**?*

```{r}
#| label: target-keyboard

rules_keyboard <- apriori(
  txns,
  parameter  = list(supp = 0.15, conf = 0.50, minlen = 2),
  appearance = list(rhs = "keyboard", default = "lhs")
)

cat("Rules predicting keyboard:", length(rules_keyboard), "\n\n")
inspect(sort(rules_keyboard, by = "lift", decreasing = TRUE))
```

## Step 10 — Export to Data Frame

```{r}
#| label: dataframe

rules_df <- as(rules, "data.frame")
rules_df$support    <- round(rules_df$support,    3)
rules_df$confidence <- round(rules_df$confidence, 3)
rules_df$lift       <- round(rules_df$lift,       3)

head(rules_df[order(-rules_df$lift), ], 8)
```

## Summary of Four Measures {background-color="#2c3e50"}

**Rule: {laptop, mouse} → {keyboard}**

| Measure        | Value | Meaning                             |
|----------------|-------|-------------------------------------|
| **Support**    | 0.40  | Appears in 40% of transactions      |
| **Confidence** | 0.80  | 80% of the time rule is correct     |
| **Lift**       | 1.00  | Independent — no better than chance  |
| **Conviction** | 1.00  | Confirms independence               |

**Rule: {mouse} → {mousepad}**

| Measure        | Value | Meaning                              |
|----------------|-------|--------------------------------------|
| **Support**    | 0.30  | Appears in 30% of transactions       |
| **Confidence** | 0.429 | 43% of mouse buyers also buy mousepad|
| **Lift**       | 1.43  | 43% more likely than random          |
| **Conviction** | 1.23  | Genuine dependency                   |

## Key Takeaways {background-color="#2c3e50"}

::: incremental
-   **Support** — how *common* is the pattern?
-   **Confidence** — how *reliable* is the rule?
-   **Lift > 1** — the rule is *better* than random guessing
-   **Conviction > 1** — items are *genuinely dependent*
-   Always check **Lift** before trusting a high-confidence rule
-   The **Apriori algorithm** prunes efficiently using the anti-monotone property
-   `arules` + `arulesViz` make ARM easy in R
:::

. . .

::: callout-tip
## Try It Yourself!

1.  Change `supp` and `conf` thresholds — how does the rule count change?
2.  Add more transactions and items — do the measures improve?
3.  Filter for `"mouse"` on the RHS — what predicts mouse purchases?
:::

## References

1.  Agrawal, R., & Srikant, R. (1994). *Fast algorithms for mining association rules*. VLDB.
2.  Hahsler, M., Grün, B., & Hornik, K. (2005). `arules` — A computational environment for mining association rules. *Journal of Statistical Software*, 14(15).
3.  Hahsler, M. (2017). `arulesViz`: Visualizing Association Rules and Frequent Itemsets. *Journal of Statistical Software*, 76(2).
4.  Brin, S., Motwani, R., Ullman, J., & Tsur, S. (1997). *Dynamic itemset counting and implication rules for market basket data*. ACM SIGMOD.
