---
title: "Exploratory Data Analysis on mtcars — Lecture Handout"
subtitle: "Basic-Level EDA with R"
author: "Dr. Mustafa Hameed"
date: today
format:
  pdf:
    documentclass: article
    papersize: letter
    geometry:
      - margin=1in
    fontsize: 11pt
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    linkcolor: blue
    highlight-style: github
    header-includes: |
      \usepackage{booktabs}
      \usepackage{longtable}
      \usepackage{colortbl}
      \usepackage{xcolor}
      \definecolor{darkblue}{HTML}{2C3E50}
      \definecolor{accentblue}{HTML}{2980B9}
      \usepackage{fancyhdr}
      \pagestyle{fancy}
      \fancyhead[L]{EDA Lecture Handout — mtcars}
      \fancyhead[R]{Dr. Mustafa Hameed}
      \fancyfoot[C]{\thepage}
execute:
  echo: true
  warning: false
  message: false
---

# What is Exploratory Data Analysis (EDA)?

**Exploratory Data Analysis** is the process of examining and summarizing a dataset to understand its main characteristics *before* applying any formal statistical models or machine learning algorithms.

**EDA is the FIRST step in every data science project.**

## Why Do We Do EDA?

| Purpose | What You Learn |
|:--------|:---------------|
| **Inspect structure** | How many rows/columns? What are the variable types? |
| **Detect errors** | Missing values, impossible numbers, wrong data types |
| **Summarize** | Mean, median, range, standard deviation |
| **Visualize** | Distributions, relationships, outliers |
| **Guide modeling** | Which variables matter? What transformations are needed? |

## Key Statistical Concepts

| Term | Definition |
|:-----|:-----------|
| **Mean** | The average: sum of all values ÷ count |
| **Median** | The middle value when data is sorted |
| **Standard Deviation (SD)** | Measures how spread out values are around the mean |
| **Quartiles** | Split data into 4 equal parts (Q1 = 25%, Q2 = 50%, Q3 = 75%) |
| **Correlation** | Measures the strength of linear relationship between two variables (−1 to +1) |
| **Outlier** | A value that is unusually far from the rest of the data |

---

# The mtcars Dataset

The `mtcars` dataset is built into R. It comes from the 1974 *Motor Trend* US magazine and contains data on 32 automobiles (1973–74 models).

## Variable Descriptions

| Variable | Description | Type |
|:---------|:-----------|:-----|
| `mpg` | Miles per gallon (fuel efficiency) | Continuous |
| `cyl` | Number of cylinders (4, 6, 8) | Categorical |
| `disp` | Displacement in cubic inches | Continuous |
| `hp` | Gross horsepower | Continuous |
| `drat` | Rear axle ratio | Continuous |
| `wt` | Weight (× 1000 lbs) | Continuous |
| `qsec` | Quarter-mile time (seconds) | Continuous |
| `vs` | Engine shape (0 = V-shaped, 1 = straight) | Binary |
| `am` | Transmission (0 = automatic, 1 = manual) | Binary |
| `gear` | Number of forward gears (3, 4, 5) | Categorical |
| `carb` | Number of carburetors | Categorical |

---

# EDA Workflow — Step by Step

## Step 1: Load and Inspect

```{r}
# Dimensions
dim(mtcars)

# First 6 rows
head(mtcars)

# Structure
str(mtcars)
```

**What to look for:** 32 rows × 11 columns. All numeric types. Some variables (cyl, vs, am, gear, carb) are actually categorical.

## Step 2: Summary Statistics

```{r}
summary(mtcars)
```

**Key observations:**

- MPG ranges from **10.4 to 33.9** (mean = 20.09)
- HP ranges from **52 to 335** (mean = 146.7)
- Weight ranges from **1.51 to 5.42** thousand lbs (mean = 3.22)

## Step 3: Frequency Tables

```{r}
# How many cars per cylinder group?
table(mtcars$cyl)

# Transmission breakdown
table(mtcars$am)
```

**Findings:** 14 cars have 8 cylinders, 11 have 4, and 7 have 6. 19 are automatic, 13 manual.

## Step 4: Visualize Distributions

### Histogram — MPG Distribution

```{r}
#| fig-width: 5
#| fig-height: 3.5
hist(mtcars$mpg, col = "steelblue", border = "white",
     main = "Distribution of MPG", xlab = "Miles Per Gallon",
     ylab = "Frequency", breaks = 10)
abline(v = mean(mtcars$mpg), col = "red", lwd = 2, lty = 2)
```

**Interpretation:** Most cars get 15–22 MPG. The distribution is slightly right-skewed.

### Boxplot — MPG by Cylinders

```{r}
#| fig-width: 5
#| fig-height: 3.5
boxplot(mpg ~ cyl, data = mtcars,
        col = c("lightgreen", "gold", "salmon"),
        main = "MPG by Cylinders", xlab = "Cylinders", ylab = "MPG")
```

**Interpretation:** 4-cylinder cars are the most fuel-efficient. 8-cylinder cars are the least.

### Boxplot — MPG by Transmission

```{r}
#| fig-width: 5
#| fig-height: 3.5
boxplot(mpg ~ am, data = mtcars,
        col = c("lightblue", "lightyellow"),
        names = c("Automatic", "Manual"),
        main = "MPG by Transmission", xlab = "Transmission", ylab = "MPG")
```

**Interpretation:** Manual transmission cars tend to be more fuel-efficient.

## Step 5: Visualize Relationships

### Scatter Plot — MPG vs Weight

```{r}
#| fig-width: 5
#| fig-height: 3.5
plot(mtcars$wt, mtcars$mpg, col = "steelblue", pch = 19,
     main = "MPG vs Weight", xlab = "Weight (1000 lbs)", ylab = "MPG")
abline(lm(mpg ~ wt, data = mtcars), col = "red", lwd = 2)
```

**Interpretation:** Strong negative relationship — heavier cars get worse fuel economy.

### Scatter Plot — MPG vs Horsepower

```{r}
#| fig-width: 5
#| fig-height: 3.5
plot(mtcars$hp, mtcars$mpg, col = "coral", pch = 19,
     main = "MPG vs Horsepower", xlab = "Horsepower", ylab = "MPG")
abline(lm(mpg ~ hp, data = mtcars), col = "darkred", lwd = 2)
```

**Interpretation:** More powerful cars tend to have lower fuel efficiency.

## Step 6: Correlation Analysis

```{r}
cor_vars <- mtcars[, c("mpg", "hp", "wt", "disp", "drat", "qsec")]
round(cor(cor_vars), 2)
```

### Correlation Interpretation Guide

| Correlation Range | Strength |
|:-----------------|:---------|
| ±0.70 to ±1.00 | Strong |
| ±0.40 to ±0.69 | Moderate |
| ±0.10 to ±0.39 | Weak |
| 0.00 to ±0.09 | None |

**Key correlations:**

- mpg ↔ wt: **−0.87** (strong negative)
- mpg ↔ hp: **−0.78** (strong negative)
- mpg ↔ disp: **−0.85** (strong negative)
- wt ↔ disp: **+0.89** (strong positive)

## Step 7: Subsetting and Aggregation

```{r}
# Fuel-efficient cars (MPG > 20)
efficient <- mtcars[mtcars$mpg > 20, c("mpg", "cyl", "hp", "wt")]
efficient

# Mean MPG by cylinder group
aggregate(mpg ~ cyl, data = mtcars, FUN = mean)

# Mean MPG by transmission
agg_am <- aggregate(mpg ~ am, data = mtcars, FUN = mean)
agg_am$am <- ifelse(agg_am$am == 0, "Automatic", "Manual")
agg_am
```

---

# Summary of Key Findings

| Finding | Evidence |
|:--------|:---------|
| MPG varies widely | Range: 10.4 – 33.9 |
| Heavier cars use more fuel | Correlation: −0.87 |
| More cylinders = less efficient | 4-cyl: 26.7 MPG, 8-cyl: 15.1 MPG |
| Manual cars tend to be more efficient | Manual avg: 24.4 MPG vs Auto: 17.1 MPG |
| HP and displacement hurt fuel economy | Correlations: −0.78 and −0.85 |

---

# EDA Cheat Sheet

| Task | R Function | Example |
|:-----|:-----------|:--------|
| View data | `head()`, `tail()` | `head(mtcars)` |
| Dimensions | `dim()`, `nrow()`, `ncol()` | `dim(mtcars)` |
| Structure | `str()` | `str(mtcars)` |
| Summary stats | `summary()`, `mean()`, `sd()` | `summary(mtcars$mpg)` |
| Frequency table | `table()` | `table(mtcars$cyl)` |
| Histogram | `hist()` | `hist(mtcars$mpg)` |
| Boxplot | `boxplot()` | `boxplot(mpg ~ cyl, data = mtcars)` |
| Scatter plot | `plot()` | `plot(mtcars$wt, mtcars$mpg)` |
| Correlation | `cor()` | `cor(mtcars$mpg, mtcars$wt)` |
| Filter rows | `[ ]` or `subset()` | `mtcars[mtcars$mpg > 20, ]` |
| Aggregate | `aggregate()` | `aggregate(mpg ~ cyl, data = mtcars, mean)` |

---

# Practice Exercises

1. Create a histogram of `wt` (weight). Is it skewed?
2. Make a boxplot of `hp` grouped by `cyl`. What do you observe?
3. Calculate the correlation between `hp` and `wt`. Interpret the result.
4. Filter cars with HP > 200. How many are there? What is their average MPG?
5. Create a scatter plot of `disp` vs `mpg`. Add a regression line. Describe the trend.

---

# References

1. Henderson, H. V., & Velleman, P. F. (1981). Building multiple regression models interactively. *Biometrics*, 37, 391–411.
2. R Core Team (2024). *R: A language and environment for statistical computing*.
3. Tukey, J. W. (1977). *Exploratory Data Analysis*. Addison-Wesley.
