---
title: "Data Science Basics with R: Exploratory Data Analysis"
author: "Instructor"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

This lab introduces **Exploratory Data Analysis (EDA)** using the built-in `mtcars` dataset in R.

### Learning Objectives

1.  Understand the structure of a dataset\
2.  Perform basic exploratory analysis\
3.  Visualize data using histograms and scatter plots\
4.  Apply filtering, subsetting, and aggregation

## Viewing the Dataset

```{r}
View(mtcars)
```

## Checking Dataset Dimensions

```{r}
dim(mtcars)
```

## Previewing Data Values

```{r}
head(mtcars)
head(mtcars, n = 3)
tail(mtcars)
```

## Dataset Documentation

```{r}
?mtcars
```

## Inspecting Dataset Structure

```{r}
str(mtcars)
```

## Summary Statistics

```{r}
summary(mtcars)
```

## Histogram: Distribution of MPG

```{r}
hist(mtcars$mpg,
     col = "steelblue",
     main = "Histogram of MPG",
     xlab = "Miles Per Gallon (MPG)",
     ylab = "Frequency")
```

## Histogram: Number of Gears

```{r}
hist(mtcars$gear,
     col = "steelblue",
     main = "Histogram of Number of Gears",
     xlab = "Number of Gears",
     ylab = "Number of Cars")
```

## Scatter Plot: MPG vs Weight

```{r}
plot(mtcars$mpg, mtcars$wt,
     col = "steelblue",
     main = "Scatter Plot of MPG vs Weight",
     xlab = "Miles Per Gallon (MPG)",
     ylab = "Car Weight (1000 lbs)",
     pch = 19)
```

## Scatter Plot: MPG vs Gear

```{r}
plot(mtcars$gear, mtcars$mpg,
     col = "steelblue",
     main = "Scatter Plot of MPG vs Gear",
     xlab = "Number of Gears",
     ylab = "Miles Per Gallon (MPG)",
     pch = 19)
```

## Subsetting Data

```{r}
mt1 <- mtcars[mtcars$mpg > 20, c("gear", "mpg")]
View(mt1)
```

## Aggregation: Mean MPG by Gear

```{r}
mt2 <- aggregate(. ~ gear,
                 mtcars[mtcars$mpg > 20, c("gear", "mpg")],
                 mean)
View(mt2)
```

## Post-Aggregation Filtering

```{r}
mt3 <- subset(mt2, mpg > 25)
View(mt3)
```

## Key Takeaways

Exploratory Data Analysis is the foundation of all data science and machine learning workflows.
