---
title: "Exploratory Data Analysis on mtcars"
subtitle: "A Basic-Level EDA Tutorial with R"
author: "Dr. Mustafa Hameed"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

## What is Exploratory Data Analysis?

**Exploratory Data Analysis (EDA)** is the critical first step in any data science workflow. Before building models or running statistical tests, we must *understand* the data.

EDA involves:

- **Inspecting** — What does the data look like?
- **Summarizing** — What are the key statistics?
- **Visualizing** — What patterns and relationships exist?
- **Cleaning** — Are there missing values or outliers?

## Why EDA Matters

| Reason | Explanation |
|:-------|:-----------|
| Catch errors early | Spot missing values, wrong types, impossible values |
| Understand distributions | Is the data skewed? Are there outliers? |
| Find relationships | Which variables are correlated? |
| Guide modeling | Choose the right algorithm based on data characteristics |
| Communicate findings | Visualizations tell the story before models do |

## The mtcars Dataset

The `mtcars` dataset is built into R. It contains data from the 1974 *Motor Trend* US magazine, covering fuel consumption and 10 aspects of automobile design and performance for 32 cars.

### Variable Definitions

| Variable | Description | Type |
|:---------|:-----------|:-----|
| `mpg` | Miles per gallon (fuel efficiency) | Continuous |
| `cyl` | Number of cylinders (4, 6, 8) | Categorical |
| `disp` | Displacement (cubic inches) | Continuous |
| `hp` | Gross horsepower | Continuous |
| `drat` | Rear axle ratio | Continuous |
| `wt` | Weight (1000 lbs) | Continuous |
| `qsec` | Quarter-mile time (seconds) | Continuous |
| `vs` | Engine shape (0 = V-shaped, 1 = straight) | Binary |
| `am` | Transmission (0 = automatic, 1 = manual) | Binary |
| `gear` | Number of forward gears (3, 4, 5) | Categorical |
| `carb` | Number of carburetors | Categorical |

---

# Step 1: Load and Inspect the Data

## View the Dataset

```{r}
# Display first 6 rows (default)
head(mtcars)
```

## Check Dimensions

```{r}
# How many rows (cars) and columns (variables)?
dim(mtcars)
cat("Rows:", nrow(mtcars), " Columns:", ncol(mtcars), "\n")
```

**Interpretation:** We have **32 observations** (cars) and **11 variables** (features).

## Preview Top and Bottom Rows

```{r}
# First 5 rows
head(mtcars, n = 5)

# Last 5 rows
tail(mtcars, n = 5)
```

## Inspect Structure

```{r}
# Variable names, types, and sample values
str(mtcars)
```

**Interpretation:** All variables are stored as `numeric` (type `num`). Some should be treated as **categorical** for analysis (e.g., `cyl`, `vs`, `am`, `gear`, `carb`).

## Column Names

```{r}
names(mtcars)
```

---

# Step 2: Summary Statistics

## Descriptive Statistics

```{r}
summary(mtcars)
```

**How to read this output:**

| Statistic | Meaning |
|:----------|:--------|
| Min | Smallest value |
| 1st Qu. | 25th percentile — 25% of data falls below this |
| Median | Middle value (50th percentile) |
| Mean | Average |
| 3rd Qu. | 75th percentile — 75% of data falls below this |
| Max | Largest value |

## Key Observations from Summary

```{r}
cat("MPG range:", min(mtcars$mpg), "to", max(mtcars$mpg), "\n")
cat("Average MPG:", round(mean(mtcars$mpg), 2), "\n")
cat("Median MPG:", median(mtcars$mpg), "\n")
cat("Average HP:", round(mean(mtcars$hp), 2), "\n")
cat("Average Weight:", round(mean(mtcars$wt), 2), "× 1000 lbs\n")
```

## Standard Deviation

```{r}
# Standard deviation measures spread around the mean
cat("SD of MPG:", round(sd(mtcars$mpg), 2), "\n")
cat("SD of HP:", round(sd(mtcars$hp), 2), "\n")
cat("SD of Weight:", round(sd(mtcars$wt), 2), "\n")
```

**Interpretation:** A larger SD means more variability. HP has a large SD (~69), meaning horsepower varies widely across these cars.

---

# Step 3: Frequency Counts for Categorical Variables

```{r}
cat("Cylinders:\n")
table(mtcars$cyl)

cat("\nTransmission (0=auto, 1=manual):\n")
table(mtcars$am)

cat("\nGears:\n")
table(mtcars$gear)

cat("\nEngine shape (0=V, 1=straight):\n")
table(mtcars$vs)
```

**Interpretation:**

- Most cars have **8 cylinders** (14 cars), followed by 4 (11) and 6 (7).
- **19 cars** are automatic, **13** are manual.
- Most cars have **3 gears** (15), then 4 (12), then 5 (5).

---

# Step 4: Visualizations

## 4.1 Histogram — Distribution of MPG

```{r}
#| label: hist-mpg
#| fig-cap: "Distribution of fuel efficiency (MPG)"
hist(mtcars$mpg,
     col = "steelblue",
     border = "white",
     main = "Histogram of MPG",
     xlab = "Miles Per Gallon",
     ylab = "Frequency",
     breaks = 10)
abline(v = mean(mtcars$mpg), col = "red", lwd = 2, lty = 2)
legend("topright", legend = paste("Mean =", round(mean(mtcars$mpg), 1)),
       col = "red", lty = 2, lwd = 2)
```

**Interpretation:** The distribution of MPG is roughly **right-skewed** — most cars cluster around 15–22 MPG, with a few highly efficient outliers above 30 MPG. The red dashed line shows the mean.

## 4.2 Histogram — Horsepower

```{r}
#| label: hist-hp
#| fig-cap: "Distribution of horsepower"
hist(mtcars$hp,
     col = "coral",
     border = "white",
     main = "Histogram of Horsepower",
     xlab = "Gross Horsepower",
     ylab = "Frequency",
     breaks = 10)
abline(v = mean(mtcars$hp), col = "darkred", lwd = 2, lty = 2)
```

**Interpretation:** Horsepower is also **right-skewed**, with most cars between 50–175 HP and a few high-performance outliers above 250 HP.

## 4.3 Boxplot — MPG by Number of Cylinders

```{r}
#| label: box-cyl
#| fig-cap: "MPG grouped by number of cylinders"
boxplot(mpg ~ cyl, data = mtcars,
        col = c("lightgreen", "gold", "salmon"),
        main = "MPG by Number of Cylinders",
        xlab = "Cylinders",
        ylab = "Miles Per Gallon")
```

**Interpretation:**

- **4-cylinder** cars have the highest MPG (most fuel-efficient).
- **8-cylinder** cars have the lowest MPG (least fuel-efficient).
- There is a clear **negative relationship** between cylinders and fuel efficiency.

## 4.4 Boxplot — MPG by Transmission Type

```{r}
#| label: box-am
#| fig-cap: "MPG grouped by transmission type"
boxplot(mpg ~ am, data = mtcars,
        col = c("lightblue", "lightyellow"),
        names = c("Automatic", "Manual"),
        main = "MPG by Transmission Type",
        xlab = "Transmission",
        ylab = "Miles Per Gallon")
```

**Interpretation:** Manual transmission cars tend to have **higher MPG** than automatic cars in this dataset.

## 4.5 Scatter Plot — MPG vs Weight

```{r}
#| label: scatter-wt
#| fig-cap: "Scatter plot of fuel efficiency vs car weight"
plot(mtcars$wt, mtcars$mpg,
     col = "steelblue",
     pch = 19,
     main = "MPG vs Weight",
     xlab = "Weight (1000 lbs)",
     ylab = "Miles Per Gallon")
abline(lm(mpg ~ wt, data = mtcars), col = "red", lwd = 2)
```

**Interpretation:** There is a strong **negative linear relationship** — heavier cars get worse fuel economy. The red line is the linear regression fit.

## 4.6 Scatter Plot — MPG vs Horsepower

```{r}
#| label: scatter-hp
#| fig-cap: "Scatter plot of fuel efficiency vs horsepower"
plot(mtcars$hp, mtcars$mpg,
     col = "coral",
     pch = 19,
     main = "MPG vs Horsepower",
     xlab = "Horsepower",
     ylab = "Miles Per Gallon")
abline(lm(mpg ~ hp, data = mtcars), col = "darkred", lwd = 2)
```

**Interpretation:** More powerful cars tend to have **lower fuel efficiency** — a strong negative relationship.

## 4.7 Bar Chart — Number of Cars by Cylinders

```{r}
#| label: bar-cyl
#| fig-cap: "Count of cars by cylinder group"
barplot(table(mtcars$cyl),
        col = c("lightgreen", "gold", "salmon"),
        main = "Number of Cars by Cylinder Count",
        xlab = "Cylinders",
        ylab = "Count")
```

## 4.8 Pairs Plot — Key Variables

```{r}
#| label: pairs-plot
#| fig-cap: "Pairs plot of key numeric variables"
pairs(mtcars[, c("mpg", "hp", "wt", "disp", "qsec")],
      col = "steelblue",
      pch = 19,
      main = "Pairs Plot — Key Variables")
```

**Interpretation:** The pairs plot shows scatter plots for every pair of selected variables. Look for strong linear trends (e.g., mpg vs wt) and weak relationships (e.g., qsec vs disp).

---

# Step 5: Correlation Analysis

```{r}
# Correlation matrix for key variables
cor_matrix <- round(cor(mtcars[, c("mpg", "hp", "wt", "disp", "drat", "qsec")]), 2)
cor_matrix
```

**How to read correlations:**

| Value | Meaning |
|:------|:--------|
| +1.00 | Perfect positive relationship |
| +0.50 to +0.99 | Strong positive |
| +0.10 to +0.49 | Weak positive |
| 0.00 | No relationship |
| −0.10 to −0.49 | Weak negative |
| −0.50 to −0.99 | Strong negative |
| −1.00 | Perfect negative relationship |

**Key findings:**

- **mpg vs wt:** Strong negative (r ≈ −0.87) — heavier cars use more fuel
- **mpg vs hp:** Strong negative (r ≈ −0.78) — more powerful cars use more fuel
- **mpg vs disp:** Strong negative (r ≈ −0.85) — larger engines use more fuel
- **wt vs disp:** Strong positive (r ≈ 0.89) — heavier cars have bigger engines

---

# Step 6: Subsetting and Filtering

## Filter: Fuel-Efficient Cars (MPG > 20)

```{r}
efficient_cars <- mtcars[mtcars$mpg > 20, c("mpg", "cyl", "hp", "wt", "am")]
efficient_cars
cat("\nNumber of fuel-efficient cars:", nrow(efficient_cars), "out of 32\n")
```

## Aggregate: Mean MPG by Cylinders

```{r}
agg_cyl <- aggregate(mpg ~ cyl, data = mtcars, FUN = mean)
names(agg_cyl) <- c("Cylinders", "Mean_MPG")
agg_cyl
```

**Interpretation:** 4-cylinder cars average ~26.7 MPG, 6-cylinder ~19.7, and 8-cylinder ~15.1.

## Aggregate: Mean MPG by Transmission

```{r}
agg_am <- aggregate(mpg ~ am, data = mtcars, FUN = mean)
agg_am$am <- ifelse(agg_am$am == 0, "Automatic", "Manual")
names(agg_am) <- c("Transmission", "Mean_MPG")
agg_am
```

## Aggregate: Mean MPG by Gear Count

```{r}
agg_gear <- aggregate(mpg ~ gear, data = mtcars, FUN = mean)
names(agg_gear) <- c("Gears", "Mean_MPG")
agg_gear
```

---

# Step 7: Advanced Filtering

```{r}
# Cars with MPG > 25 AND manual transmission
high_eff_manual <- mtcars[mtcars$mpg > 25 & mtcars$am == 1,
                          c("mpg", "cyl", "hp", "wt", "gear")]
cat("High-efficiency manual cars:\n")
high_eff_manual
```

```{r}
# Cars with 4 cylinders, sorted by MPG (descending)
four_cyl <- mtcars[mtcars$cyl == 4, ]
four_cyl <- four_cyl[order(-four_cyl$mpg), c("mpg", "hp", "wt", "gear", "am")]
cat("4-cylinder cars (sorted by MPG):\n")
four_cyl
```

---

# Summary

## What We Learned

| EDA Step | What It Does | Key R Functions |
|:---------|:------------|:----------------|
| Inspect | See the data structure | `head()`, `str()`, `dim()` |
| Summarize | Compute statistics | `summary()`, `mean()`, `sd()`, `table()` |
| Visualize | See patterns | `hist()`, `boxplot()`, `plot()`, `pairs()` |
| Correlate | Measure relationships | `cor()` |
| Filter | Focus on subsets | `[ ]`, `subset()`, `aggregate()` |

## Key Findings from mtcars

1. **MPG ranges from 10.4 to 33.9** — wide variation in fuel efficiency
2. **Heavier cars have worse fuel economy** (r ≈ −0.87)
3. **More cylinders = less fuel efficient** — 4-cyl averages 26.7 MPG vs 8-cyl at 15.1
4. **Manual transmission cars** tend to have higher MPG than automatics
5. **HP, weight, and displacement** are the strongest predictors of fuel efficiency

## Next Steps

- Apply **ggplot2** for more polished visualizations
- Perform **hypothesis testing** (e.g., is the MPG difference by transmission significant?)
- Build a **linear regression model** to predict MPG
- Explore **feature engineering** (e.g., power-to-weight ratio)

---

# References

1. Henderson, H. V., & Velleman, P. F. (1981). Building multiple regression models interactively. *Biometrics*, 37, 391–411.
2. R Core Team (2024). R: A language and environment for statistical computing.
3. Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*. Springer.
