---
title: "Predictive Modeling with Machine Learning in R"
subtitle: "Predicting Student Performance from LMS Engagement Data"
author: "Dr. Mustafa Hameed"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo
    self-contained: true
---

## Introduction

### What is Predictive Modeling?

**Predictive modeling** uses historical data to build mathematical models that can predict future or unknown outcomes. In education, this means using data about student behavior to predict their academic performance.

### Why Predictive Modeling Matters

- **Early Intervention**: Identify at-risk students before they fail
- **Resource Allocation**: Direct support to students who need it most
- **Understanding Behavior**: Learn which engagement patterns lead to success
- **Data-Driven Decisions**: Move beyond intuition to evidence-based strategies

### Our Goal

We will predict **Final_Grade** (a continuous numeric outcome) using student engagement features from a Learning Management System (LMS). This is a **regression** problem.

### Machine Learning Workflow

```
1. Load Data → 2. Explore (EDA) → 3. Prepare Data → 4. Split (Train/Test)
→ 5. Build Models → 6. Evaluate & Compare → 7. Interpret Results
```

---

## Setup

### Load Required Libraries

```{r}
#| message: false
#| warning: false

# Data manipulation and visualization
library(tidyverse)

# Data import
library(rio)

# Summary statistics
library(skimr)

# Correlation analysis
library(correlation)

# Random Forest
library(randomForest)

# Train/Test splitting
library(rsample)

# Model evaluation metrics
library(yardstick)

# Model performance checks
library(performance)
```

### Load the Dataset

We use a synthetic dataset of 285 students with LMS engagement metrics.

```{r}
# Load data from GitHub
lms <- rio::import("https://github.com/lamethods/data2/raw/main/lms/lms.csv")

# Quick look at the data
glimpse(lms)
```

### Understanding the Variables

| Variable | Description |
|----------|-------------|
| `Freq_Course_View` | How often student views course pages |
| `Freq_Lecture_View` | How often student views lectures |
| `Freq_Forum_Consume` | How often student reads forum posts |
| `Freq_Forum_Contribute` | How often student writes forum posts |
| `Regularity_Course_View` | Consistency of course page visits |
| `Regularity_Lecture_View` | Consistency of lecture viewing |
| `Regularity_Forum_Consume` | Consistency of forum reading |
| `Regularity_Forum_Contribute` | Consistency of forum posting |
| `Session_Count` | Number of login sessions |
| `Total_Duration` | Total time spent on LMS |
| `Active_Days` | Number of days with LMS activity |
| **`Final_Grade`** | **Target variable — student's final grade** |

---

## Exploratory Data Analysis (EDA)

### Summary Statistics

```{r}
skim(lms)
```

### Distribution of the Target Variable

```{r}
ggplot(lms, aes(x = Final_Grade)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = mean(Final_Grade)), 
             color = "red", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Final Grade",
       subtitle = paste("Mean =", round(mean(lms$Final_Grade), 2)),
       x = "Final Grade", y = "Count") +
  theme_minimal()
```

### Correlation Analysis

```{r}
#| fig-width: 10
#| fig-height: 8

# Compute correlations
cor_matrix <- cor(lms)

# Visualize with a heatmap
cor_df <- as.data.frame(as.table(cor_matrix))
names(cor_df) <- c("Var1", "Var2", "Correlation")

ggplot(cor_df, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1, 1)) +
  geom_text(aes(label = round(Correlation, 2)), size = 2.5) +
  labs(title = "Correlation Heatmap of All Variables") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Top Correlations with Final_Grade

```{r}
# Which features correlate most with Final_Grade?
cor_with_grade <- cor(lms)[, "Final_Grade"]
cor_with_grade <- sort(abs(cor_with_grade), decreasing = TRUE)
cor_with_grade[-1]  # Remove self-correlation
```

### Scatter Plots: Key Predictors vs Final Grade

```{r}
#| fig-width: 10
#| fig-height: 6

lms_long <- lms %>%
  pivot_longer(cols = -Final_Grade, names_to = "Feature", values_to = "Value")

ggplot(lms_long, aes(x = Value, y = Final_Grade)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_wrap(~Feature, scales = "free_x", ncol = 4) +
  labs(title = "Relationship Between Each Feature and Final Grade") +
  theme_minimal()
```

---

## Data Preparation

### Standardize the Features

Standardization transforms features to have **mean = 0** and **standard deviation = 1**. This ensures all features are on the same scale.

$$z = \frac{x - \mu}{\sigma}$$

```{r}
# Separate predictors and target
predictors <- lms %>% select(-Final_Grade)
target <- lms$Final_Grade

# Standardize predictors
predictors_scaled <- as.data.frame(scale(predictors))

# Recombine
lms_scaled <- predictors_scaled %>%
  mutate(Final_Grade = target)

head(lms_scaled)
```

### Check for Multicollinearity (VIF)

**Variance Inflation Factor (VIF)** measures how much the variance of a regression coefficient is inflated due to collinearity. VIF > 5 may indicate a problem.

```{r}
# Fit a linear model to check VIF
lm_check <- lm(Final_Grade ~ ., data = lms_scaled)
check_collinearity(lm_check)
```

---

## Train/Test Split

We split the data **80% for training** and **20% for testing**. The model learns patterns from the training set and is evaluated on the unseen test set.

```{r}
set.seed(123)  # For reproducibility

# Create 80/20 split
split <- initial_split(lms_scaled, prop = 0.80)
train_data <- training(split)
test_data  <- testing(split)

cat("Training set:", nrow(train_data), "rows\n")
cat("Testing set: ", nrow(test_data), "rows\n")
```

---

## Model Building

### Model 1: Linear Regression

**Linear Regression** fits a straight-line relationship between predictors and the outcome.

$$\hat{y} = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_p x_p$$

```{r}
# Train linear regression model
model_lm <- lm(Final_Grade ~ ., data = train_data)

# Summary of model
summary(model_lm)
```

```{r}
# Make predictions on test set
pred_lm <- predict(model_lm, newdata = test_data)

# View first 10 predictions vs actual
head(data.frame(Actual = test_data$Final_Grade, Predicted = round(pred_lm, 3)), 10)
```

### Model 2: Random Forest

**Random Forest** builds many decision trees on random subsets of the data and averages their predictions. This reduces overfitting and improves accuracy.

```{r}
set.seed(123)

# Train Random Forest with 500 trees
model_rf <- randomForest(
  Final_Grade ~ ., 
  data = train_data,
  ntree = 500,
  importance = TRUE
)

# Model summary
print(model_rf)
```

```{r}
# Make predictions on test set
pred_rf <- predict(model_rf, newdata = test_data)

# View first 10 predictions vs actual
head(data.frame(Actual = test_data$Final_Grade, Predicted = round(pred_rf, 3)), 10)
```

---

## Model Evaluation

### Key Evaluation Metrics

| Metric | Formula | Interpretation |
|--------|---------|---------------|
| **RMSE** | $\sqrt{\frac{1}{n}\sum(y_i - \hat{y}_i)^2}$ | Average prediction error (lower = better) |
| **MAE** | $\frac{1}{n}\sum\|y_i - \hat{y}_i\|$ | Average absolute error (lower = better) |
| **R²** | $1 - \frac{SS_{res}}{SS_{tot}}$ | Proportion of variance explained (higher = better) |

### Calculate Metrics for Both Models

```{r}
# Create results data frames
results_lm <- data.frame(
  truth = test_data$Final_Grade,
  estimate = pred_lm
)

results_rf <- data.frame(
  truth = test_data$Final_Grade,
  estimate = pred_rf
)

# Calculate metrics for Linear Regression
cat("=== Linear Regression ===\n")
cat("RMSE:", rmse_vec(results_lm$truth, results_lm$estimate), "\n")
cat("MAE: ", mae_vec(results_lm$truth, results_lm$estimate), "\n")
cat("R²:  ", rsq_vec(results_lm$truth, results_lm$estimate), "\n\n")

# Calculate metrics for Random Forest
cat("=== Random Forest ===\n")
cat("RMSE:", rmse_vec(results_rf$truth, results_rf$estimate), "\n")
cat("MAE: ", mae_vec(results_rf$truth, results_rf$estimate), "\n")
cat("R²:  ", rsq_vec(results_rf$truth, results_rf$estimate), "\n")
```

### Comparison Table

```{r}
# Build a comparison table
comparison <- data.frame(
  Model = c("Linear Regression", "Random Forest"),
  RMSE = c(
    rmse_vec(results_lm$truth, results_lm$estimate),
    rmse_vec(results_rf$truth, results_rf$estimate)
  ),
  MAE = c(
    mae_vec(results_lm$truth, results_lm$estimate),
    mae_vec(results_rf$truth, results_rf$estimate)
  ),
  R_squared = c(
    rsq_vec(results_lm$truth, results_lm$estimate),
    rsq_vec(results_rf$truth, results_rf$estimate)
  )
)

comparison %>%
  mutate(across(where(is.numeric), round, 3)) %>%
  knitr::kable(caption = "Model Comparison: Linear Regression vs Random Forest")
```

---

## Visualizing Results

### Predicted vs Actual Values

```{r}
#| fig-width: 10
#| fig-height: 5

# Combine predictions
viz_data <- data.frame(
  Actual = rep(test_data$Final_Grade, 2),
  Predicted = c(pred_lm, pred_rf),
  Model = rep(c("Linear Regression", "Random Forest"), each = nrow(test_data))
)

ggplot(viz_data, aes(x = Actual, y = Predicted, color = Model)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~Model) +
  labs(title = "Predicted vs Actual Final Grade",
       subtitle = "Points on the dashed line = perfect predictions",
       x = "Actual Grade", y = "Predicted Grade") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Residual Plots

Residuals = Actual − Predicted. Ideally, residuals should be randomly scattered around zero.

```{r}
#| fig-width: 10
#| fig-height: 5

viz_data$Residual <- viz_data$Actual - viz_data$Predicted

ggplot(viz_data, aes(x = Predicted, y = Residual, color = Model)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~Model) +
  labs(title = "Residual Plots",
       subtitle = "Residuals should be randomly scattered around 0",
       x = "Predicted Grade", y = "Residual") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Variable Importance (Random Forest)

Variable importance shows which features have the most impact on predictions.

```{r}
#| fig-width: 8
#| fig-height: 6

# Extract importance
imp <- importance(model_rf)
imp_df <- data.frame(
  Feature = rownames(imp),
  Importance = imp[, "%IncMSE"]
)

ggplot(imp_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  coord_flip() +
  labs(title = "Random Forest: Variable Importance",
       subtitle = "% Increase in MSE when variable is permuted",
       x = "Feature", y = "Importance (% Increase in MSE)") +
  theme_minimal()
```

---

## Interpretation of Results

### Key Findings

1. **Both models can predict student grades** from LMS engagement data, explaining a meaningful portion of the variance.

2. **Linear Regression** provides interpretable coefficients — you can see exactly how each feature contributes to the prediction.

3. **Random Forest** captures non-linear relationships that linear regression may miss, often producing better accuracy.

4. **Most Important Features** typically include:
   - **Session_Count**: Number of login sessions
   - **Active_Days**: Days with LMS activity
   - **Total_Duration**: Time spent on the platform
   - **Freq_Lecture_View**: How often lectures are viewed

### Practical Implications

- Students who **log in more frequently** and **spend more time** on the LMS tend to score higher.
- **Regular engagement** (not just total amount) matters for predicting success.
- These models can power **early warning systems** in educational settings.

---

## Summary

### What We Learned

| Step | What We Did |
|------|-------------|
| **1. Load Data** | Imported 285 student records with 12 variables |
| **2. EDA** | Explored distributions, correlations, and patterns |
| **3. Prepare** | Standardized features for fair comparison |
| **4. Split** | 80% training / 20% testing |
| **5. Build** | Trained Linear Regression and Random Forest |
| **6. Evaluate** | Compared using RMSE, MAE, and R² |
| **7. Interpret** | Identified key engagement features |

### Key Definitions

- **Supervised Learning**: Learning from labeled examples (we know the outcome)
- **Regression**: Predicting a continuous numeric value
- **Training Set**: Data used to teach the model
- **Test Set**: Held-out data used to evaluate the model
- **Overfitting**: When a model learns noise instead of patterns
- **RMSE**: Root Mean Squared Error — penalizes large errors
- **MAE**: Mean Absolute Error — average error magnitude
- **R²**: Coefficient of Determination — proportion of variance explained

---

## Practice Exercises

1. **Change the split ratio** to 70/30 instead of 80/20. How do the results change?
2. **Try removing** the least important variable. Does performance improve?
3. **Increase the number of trees** in Random Forest to 1000. Does R² improve?
4. **Plot** the histogram of residuals for each model. Are they normally distributed?
5. **Research**: What is cross-validation and why might it be better than a single train/test split?

---

*Dataset source: [lamethods/data2](https://github.com/lamethods/data2) — Synthetic LMS engagement data*
